---
title: '{ggtext} for images as x-Axis'
author: Jonathan Carroll
date: '2019-08-13'
slug: ggtext-for-images-as-x-axis
categories:
  - R
tags:
  - ggplot2
  - rstats
type: ''
subtitle: ''
image: ''
---

```{r, setup, include = FALSE}
library(ggplot2)

knitr::opts_chunk$set(
  class.output  = "bg-success",
  class.message = "bg-info text-info",
  class.warning = "bg-warning text-warning",
  class.error   = "bg-danger text-danger"
)
```

I've written [a](https://jcarroll.com.au/2016/06/02/images-as-x-axis-labels/)
[few](https://jcarroll.com.au/2016/06/03/images-as-x-axis-labels-updated/)
[times](https://jcarroll.com.au/2018/10/16/even-more-images-as-x-axis-labels/)
about using an image as an x-axis label, and the solutions have been slowly
improving. This one blows all of them out of the water.

<!--more-->

I've written [a](https://jcarroll.com.au/2016/06/02/images-as-x-axis-labels/)
[few](https://jcarroll.com.au/2016/06/03/images-as-x-axis-labels-updated/)
[times](https://jcarroll.com.au/2018/10/16/even-more-images-as-x-axis-labels/)
about using an image as an x-axis label, and the solutions have been slowly
improving. This one blows all of them out of the water.

[Claus Wilke](https://github.com/clauswilke) ([\@ClausWilke](https://twitter.com/ClausWilke)) 
now has a [{ggtext}](https://github.com/clauswilke/ggtext) package which can very 
neatly add images as x-axis labels!

This makes the solution as simple as

```{r, fig.cap = "{ggtext} makes this easy!"}
library(ggplot2)
library(ggtext)
library(rvest)

## GDP per capita, top 11 countries
url      <- "https://en.wikipedia.org/wiki/List_of_countries_by_GDP_(nominal)_per_capita"
html     <- xml2::read_html(url)
gdppc    <- html_table(html_nodes(html, "table")[3])[[1]][1:11,]

## clean up; remove non-ASCII and perform type conversions
gdppc$Country <- gsub("Â ", "", gdppc$Country)
gdppc$Country <- iconv(gdppc$Country, "latin1", "ASCII", sub="")
gdppc$Country[9] <- "United States of America"
gdppc$Rank    <- iconv(gdppc$Rank,    "latin1", "ASCII", sub="")
gdppc$`US$`   <- as.integer(sub(",", "", gdppc$`US$`))

## switching to a different source of flag images
labels <- setNames(
  paste0("<img src='http://www.senojflags.com/images/country-flag-icons/", 
         gsub(" ", "-", gdppc$Country), "-Flag.png'  width='30' /><br>", 
         sapply(strwrap(gdppc$Country, width = 10, simplify = FALSE), 
                function(x) paste(x, collapse = "<br>"))),
  gdppc$Country
)

## create a dummy dataset
npoints <- length(gdppc$Country)
y       <- gdppc$`US$`
x       <- gdppc$Country
dat     <- data.frame(x=factor(x, levels=gdppc$Country), y=y)

## NB: #85bb65 is the color of money in the USA apparently.
gg <- ggplot(dat, aes(x=x, y=y/1e3L, group=1)) 
gg <- gg + geom_bar(col="black", fill="#85bb65", stat="identity")
gg <- gg + scale_x_discrete(name = NULL, labels = labels)
gg <- gg + theme_minimal()
gg <- gg + scale_fill_discrete(guide=FALSE)
gg <- gg + theme(plot.background = element_rect(fill="grey90"))
gg <- gg + labs(title="GDP per capita", 
                subtitle="Top 11 countries", 
                x="", y="$US/1000", 
                caption=paste0("Source: ",url))
## ggtext::element_markdown
gg <- gg + theme(axis.text.x  = element_markdown(color = "black", size = 7), 
                 axis.text.y  = element_text(size=14),
                 axis.title.x = element_blank())
gg
```

This is an even nicer solution than I've been using, not only because it's shorter 
to use, but also more flexible - markdown can be processed (using `element_markdown`) 
wherever `element_text` is used, such as axis text, legends, titles, etc...

I think this finally closes this chapter, but now it's time to make some really cool 
graphs with images.

<br />
<details>
  <summary>
    <tt>devtools::session_info()</tt>
  </summary>
```{r sessionInfo, echo = FALSE}
devtools::session_info()
```
</details>
<br />
