---
title: Polyglot Exploration of Function Overloading
author: Jonathan Carroll
date: '2023-04-03'
categories:
  - fortran
  - haskell
  - julia
  - R
  - rust
tags:
  - fortran
  - haskell
  - julia
  - R
  - rust
slug: polyglot-overloading
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
  class.output  = "bg-success",
  class.message = "bg-info text-info",
  class.warning = "bg-warning text-warning",
  class.error   = "bg-danger text-danger"
)
```

I've been working my way through [Exercism][exercism] exercises in a variety of 
languages because I strongly believe every language you learn something about teaches 
you about all the others you know, and makes for useful comparisons between what 
features they offer. I was^[in part due to a strong representation of 
Haskell at my local Functional Programming Meetup] Learning Me a [Haskell][exercism_haskell] for Great Good 
(there's [a guide/book][lyahfgg] by that name) and something about [Pattern Matching][patternmatching]
just seemed extremely familiar. 

<!--more-->

I've been working my way through [Exercism][exercism] exercises in a variety of 
languages because I strongly believe every language you learn something about teaches 
you about all the others you know, and makes for useful comparisons between what 
features they offer. I was^[in part due to a strong representation of 
Haskell at my local Functional Programming Meetup] Learning Me a [Haskell][exercism_haskell] for Great Good 
(there's [a guide/book][lyahfgg] by that name) and something about [Pattern Matching][patternmatching]
just seemed extremely familiar. 

Pattern Matching is sort of like a `case` statement, but rather than just comparing literal 
values against some `enum`, it takes into consideration how the input _"looks"_. A simple example 
is to match against either an empty list `[]` (just that; an empty list) or a non-empty list denoted 
`(x:xs)`. In Haskell, `:` is a concatenation operator (`cons` in lisp) so this is the concatenation 
of `x` and the rest of a list, `xs`. The wildcard pattern `_` matching "whatever". 

A `map` function definition (from [here][patternmatching]) is then

```haskell
map _ []     = []
map f (x:xs) = f x : map f xs
```

This is two definitions for `map`, depending on which pattern is provided as the two arguments. The first 
takes "whatever" (doesn't matter, is ignored) and an empty list and just returns an empty list. The 
second takes some function `f` and a non-empty list, and concatenates (`:`) `(f x)` (the first 
element of the list `x` provided to the function `f`) with `map f xs` (the result of providing `f` and the 
rest of the list `xs` to `map`, recursively).

Since Haskell is strongly typed, I don't think this can be used to define the same named function for 
different _types_, but it can certainly do something different depending on the pattern of the data. 
In this example, if the argument is an empty list, return `0`; if the argument is a length-1 list (`arg1` 
concatenated with an empty list) then return `arg1 * 100`, and if the argument is a longer list, return 
the product of the first element and the second. This then prints out calling `fun 5.0` and `fun [5.0, 5.0]`

```haskell
fun :: [Float] -> Float
fun [] = 0.0
fun (arg1:[]) = arg1 * 100.0
fun (arg1:arg2) = arg1 * (head arg2)

main = do
  print (fun [5.0])
  print (fun [5.0, 5.0])
```
```{R, class.source = "bg-success", eval = FALSE}
500.0
25.0
```

Woo! A different function called depending on the input. I believe it might be possible to actually 
have optional arguments via the `Data.Maybe` package but I couldn't get it to compile an example the way 
I wanted^[I'm highly likely doing something wrong - I never wrote any Haskell before last week].

Rust has [something similar][rustpattern] but more specific to a `case` statement; a `match` expression 
can take patterns as options and return whichever matches (example from [here][comprehensive])

```
fn main() {
    let input = 's';

    match input {
        'q'                   => println!("Quitting"),
        'a' | 's' | 'w' | 'd' => println!("Moving around"),
        '0'..='9'             => println!("Number input"),
        _                     => println!("Something else"),
    }
}
```
```{R, class.source = "bg-success", eval = FALSE}
Moving around
```

Another common use of `match` is to switch between the `enums` `Some` and `None` 
or `Ok` and `Err` (see [here][rustenum]).

The familiarity of the Haskell pattern matching / function definition took me back to one of the
very first programming 'tricks' I learned way back in the late 2000's working on my PhD, using Fortran;
"function overloading". I wasn't formally taught programming at all (an oversight, given how important 
it became to doing my research), so I just had to pick up bits and pieces from people who knew more.

I had a bunch of integration routines^[[Numerical Recipes in Fortran 90][nrbook] was about the most 
important book we had for writing code, basically nothing else was trusted - getting a *digital* copy 
of the code was considered a sign of true power] which were slightly different depending on whether or not
the limits were finite^[what, you don't have to integrate up to infinity in your code?], so I had to call 
the right one with various `if` statements. The 'trick' I was 
taught was to use `INTERFACE / MODULE PROCEDURE` [blocks][modproblock] to "dispatch" depending on the function 
signature, or at least the number of arguments. This meant that I could just call `integrate` regardless of 
whether it was a signature with 4 arguments, or a signature with an additional argument if a bound was `Infty`.

A "small" (Fortran is hardly economical with page real-estate) example of this, 
following the Haskell example, defines two functions `Fun1arg` and `Fun2arg` which 
can be consolidated into `fun` with the `INTERFACE` block. Calling `fun(x)` or `fun(x, y)` is 
routed to the function with the relevant signature.


```fortran
MODULE exampleDispatch
  IMPLICIT NONE

  INTERFACE fun
     MODULE PROCEDURE Fun1arg, Fun2arg
  END INTERFACE fun

  CONTAINS

    ! A function that takes one argument
    ! and multiplies it by 100
    REAL FUNCTION Fun1arg(arg1)
      IMPLICIT NONE
      REAL, INTENT( IN ) :: arg1
      Fun1arg = arg1 * 100.0
    END FUNCTION Fun1arg

    ! A function that takes two arguments
    ! and multiplies them
    REAL FUNCTION Fun2arg(arg1, arg2)
      IMPLICIT NONE
      REAL, INTENT( IN ) :: arg1, arg2
      Fun2arg = arg1 * arg2
    END FUNCTION Fun2arg

END MODULE exampleDispatch

PROGRAM dispatch

  USE exampleDispatch

  IMPLICIT NONE
  REAL :: a = 5.0
  REAL :: fun

  PRINT *, fun(a)
  PRINT *, fun(a, a)

END PROGRAM dispatch
```
```{R, class.source = "bg-success", eval = FALSE}
   500.000000    
   25.0000000
```

That takes me back! I'm going to dig out my old research code and get it into GitHub for 
posterity. I'm also going to do the [Fortran exercises in Exercism][exercism_fortran] to 
reminisce some more.

So, not quite the same as the Haskell version, but it got me thinking about dispatch. R has 
several approaches. The most common is [S3][advrS3] in which dispatch occurs based on the `class` 
of the first argument to a function, so you can have something different happen to a `data.frame` 
argument and a `tibble` argument, but in both cases the signature has the same "shape" - only the 
_types_ vary. 

Wiring that up to work differently with a `list` and any other value (the default case, which 
would break for anything that doesn't vectorize, but it's a toy example) looks like

```r
fun <- function(x) {
  UseMethod("fun")
}

fun.default <- function(x) { 
  x * 100
}

fun.list <- function(x) {
  x[[1]] * x[[2]]
}

fun(5)
fun(list(5, 5))
```
```{R, class.source = "bg-success", eval = FALSE}
[1] 500
[1] 25
```

Another option is to use [S4][advrS4] which is more complicated but more powerful. Here, dispatch 
can occur based on the entire signature, though (and I may be wrong) I believe that, too, still 
needs to have a consistent "shape". A fantastic guide to S4 is [Stuart Lee's post here][s4shortguide].

A S4 version of my example could have two options for the signature; one where both 
`x` and `y` are `"numeric"`, and another where `y` is `"missing"`. `"ANY"` would also work and 
encompass a wider scope.

```r
setGeneric("fun", function(x, y, ...) standardGeneric("fun"))

setMethod("fun", c("numeric", "missing"), function(x, y) {
  x * 100
})

setMethod("fun", c("numeric", "numeric"), function(x, y) {
  x * y
})

fun(5)
fun(5, 5)
```
```{R, class.source = "bg-success", eval = FALSE}
[1] 500
[1] 25
```

So, can we ever do what I was originally inspired to do - write a simple definition of a 
function that calculates differently depending on the number of arguments? Aha - Julia to 
the rescue!! Julia has a [beautifully simple syntax for defining methods on signatures][julia]: 
just write it out!

```julia
fun(x) = x * 100
fun(x, y) = x * y

println(fun(5))
println(fun(5, 5))
```
```{R, class.source = "bg-success", eval = FALSE}
500
25
```

That's two different signatures for `fun` computing different things, and a *lot* less 
boilerplate compared to the other languages, especially Fortran. What's written above 
is the _entire script_. You can even go further 
and be specific about the types, say, mixing `Int` and `Float64` definitions

```julia
fun(x::Int) = x * 100
fun(x::Float64) = x * 200

fun(x::Int, y::Int) = x * y
fun(x::Int, y::Float64) = x * y * 2

println(fun(5))
println(fun(5.))
println(fun(5, 5))
println(fun(5, 5.))
```
```{R, class.source = "bg-success", eval = FALSE}
500
1000.0
25
50.0
```

It doesn't get simpler or more powerful than that!!

I've added all these examples to [a repo][overloading] split out by language, and some 
instructions for running them (assuming you have the language tooling already set up).

Do you have another example from a language that does this (well? poorly?) or similar?
Leave a comment if you have one, or find me on [Mastodon](https://fosstodon.org/@jonocarroll)

<br />
<details>
  <summary>
    <tt>devtools::session_info()</tt>
  </summary>
```{r sessionInfo, echo = FALSE}
devtools::session_info()
```
</details>
<br />

[exercism]: https://exercism.org/
[lyahfgg]: https://learnyouahaskell.github.io/introduction.html
[patternmatching]: https://en.wikibooks.org/wiki/Haskell/Pattern_matching
[rustpattern]: https://doc.rust-lang.org/book/ch18-03-pattern-syntax.html
[rustenum]: https://doc.rust-lang.org/std/result/
[comprehensive]: https://google.github.io/comprehensive-rust/pattern-matching.html
[nrbook]: http://www.nrbook.com/a/bookf90pdf.html
[advrS3]: https://adv-r.hadley.nz/s3.html#s3-methods
[advrS4]: https://adv-r.hadley.nz/s4.html#s4-generics
[s4shortguide]: https://stuartlee.org/2019/07/09/s4-short-guide/
[exercism_haskell]: https://exercism.org/tracks/haskell
[exercism_fortran]: https://exercism.org/tracks/fortran
[julia]: https://docs.julialang.org/en/v1/manual/methods/
[modproblock]: https://www.intel.com/content/www/us/en/docs/fortran-compiler/developer-guide-reference/2023-0/module-procedure.html
[overloading]: https://github.com/jonocarroll/overloading