---
title: These Languages are Accumulating
author: Jonathan Carroll
date: '2024-11-28'
categories:
  - APL
  - haskell
  - julia
  - python
  - rstats
tags:
  - APL
  - haskell
  - julia
  - python
  - rstats
slug: these-languages-are-accumulating
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
  class.output  = "bg-success",
  class.message = "bg-info text-info",
  class.warning = "bg-warning text-warning",
  class.error   = "bg-danger text-danger"
)
```

I keep saying that the more programming languages you know, the more you will
understand all the others you know - I'm now at the point where I want to solve
every problem I see in a handful of different languages. They all offer
different functionality, and some are certainly more suited to particular
problems than others, but there's a world of difference between two characters
and importing from two libraries.

<!--more-->

I keep saying that the more programming languages you know, the more you will
understand all the others you know - I'm now at the point where I want to solve
every problem I see in a handful of different languages. They all offer
different functionality, and some are certainly more suited to particular
problems than others, but there's a world of difference between two characters
and importing from two libraries.

A newsletter I follow (and can't find online copies of) that demonstrates neat
things in python (gotta learn it, despite not loving it) recently covered
`accumulate`, showing that `sum` and `accumulate` were sort of related

```python
>>> my_list = [42, 73, 0, 16, 10]
>>> sum(my_list)
```

```{r, eval = FALSE, class.source = "bg-success"}
141
```

```python
>>> from itertools import accumulate
>>> list(accumulate(my_list))
```

```{r, eval = FALSE, class.source = "bg-success"}
[42, 115, 115, 131, 141]
```

`sum` adds up all the elements of the list, while `accumulate` does the same but
keeps each successive partial sum.

It rounds out the demo with an alternative function being used in `accumulate`

```python
>>> from itertools import accumulate
>>> from operator import mul  # mul(2, 3) == 6
>>> initial_investment = 1000
>>> rates = [1.01, 1.01, 1.02, 1.025, 1.035, 1.035, 1.06]
>>> list(
...     accumulate(rates, mul, initial=initial_investment)
... )
```

```{r, eval = FALSE, class.source = "bg-success"}
[1000, 1010.0, 1020.1, 1040.502, 1066.515, 1103.843, 1142.477, 1211.026]
```

Now, firstly... `from operator import mul`??? It looks like there's no way to
pass `*` as an argument to a function. I _could_ define a function that performs
the same on known arguments, e.g. `lambda x, y: x * y`

```python
>>> list(accumulate(rates, lambda x, y: x*y, initial=initial_investment))
```

```{r, eval = FALSE, class.source = "bg-success"}
[1000, 1010.0, 1020.1, 1040.502, 1066.5145499999999, 1103.8425592499998, 1142.4770488237498, 1211.0256717531747]
```

but... ew. 

It's possible that there's a different way to approach this. A list
comprehension comes to mind, e.g. something like

```python
>>> [sum(my_list[0:i]) for i in range(1, len(my_list)+1)]
```

```{r, eval = FALSE, class.source = "bg-success"}
[42, 115, 115, 131, 141]
```

but that requires performing a sum for each sub-interval, so performance would 
not scale well (admittedly, that was not a consideration here at all). I also 
don't believe there's a built-in `prod` so one must `import math` in order to do 
similar

```python
>>> import math
>>> x = [initial_investment] + rates
>>> [math.prod(x[0:i]) for i in range(1, len(x)+1)]
```

```{r, eval = FALSE, class.source = "bg-success"}
[1000, 1010.0, 1020.1, 1040.502, 1066.5145499999999, 1103.8425592499998, 1142.4770488237498, 1211.0256717531747]
```

In R that could use the built-in `cumprod` for the cumulative product

```{r}
initial_investment <- 1000
rates = c(1.01, 1.01, 1.02, 1.025, 1.035, 1.035, 1.06)

cumprod(c(initial_investment, rates))
```

but that has the 'multiply' operation hardcoded. `cumsum` uses `+` as the
function... hmmm. ~~Maybe R doesn't have a generalised `accumulate`?~~
**UPDATE** I entirely forgot that `Reduce` takes an `accumulate` argument which
does what I need here

```{r}
Reduce(`+`, 1:6)

Reduce(`+`, 1:6, accumulate = TRUE)

Reduce(`*`, 1:6)

Reduce(`*`, 1:6, accumulate = TRUE)
```

Nonetheless, building my own version is a worthwhile exercise, so the rest of 
the post remains as is.

I've been playing around with Haskell lately, so recursive functions to the
rescue! One feature of recursive functions in R that I really like is `Recall`
which calls the function _in which it is defined_ with a new set of arguments -
perfect for recursion!

```{r}
accumulate_recall <- function(x, f, i=x[1]) {
  if (!length(x)) return(NULL)
  c(i, Recall(tail(x, -1), f, f(i, x[2])))
}
```

It's also robust against renaming the function; the body doesn't actually call
`accumulate_recall` by name at all.

This might be inefficient, though - it's not uncommon to blow out the stack, so
a new `Tailcall` function (which doesn't have the same elegance of being robust
against renaming) helps with flagging this as something that can be optimised

```{r}
accumulate <- function(x, f, i=x[1]) {
  if (!length(x)) return(NULL)
  c(i, Tailcall(accumulate, tail(x, -1), f, f(i, x[2])))
}
```

With this, I can emulate the `cumsum()` and `cumprod()` functions

```{r}
cumprod(1:6)

accumulate(1:6, `*`)

cumsum(2:6)

accumulate(2:6, `+`)
```

unless I try to calculate something too big...

```{r, warning = TRUE}
cumprod(5:15)

accumulate(5:15, `*`)
```

It appears that the built-in functions convert to numeric. That's easily fixed
on input

```{r}
accumulate(as.numeric(5:15), `*`)
```

In any case, there's a generalised `accumulate` that takes the bare functions as
arguments.

But it can be so much cleaner than this!

In APL you won't find any function named "sum" because it is just a reduction
(`Reduce` in R) with the function `+`

```apl
      sum←+/
      
      sum ⍳6 ⍝ sum the values 1:6
21

      sum 1↓⍳6 ⍝ sum the values 2:6
20
```

which in R is

```{r}
sum(1:6)

sum(2:6)
```

Why would you write `sum` if you can just use `+/`? It's fewer
characters to write out the implementation than the name!

For `accumulate` the terminology in APL is `scan` which uses a very similar
glyph because the operation itself is very similar; a `reduce` (`/`) is just the
last value of a `scan` (`\`) which keeps the progressive values. In both cases,
the operator (either slash) takes a binary function as the left argument and
produces a modified function - in these examples, effectively `sum` and `prod` -
which is then applied to values on the right. The `scan` version does the same

```apl
      +\⍳6
1 3 6 10 15 21

      ×\⍳6
1 2 6 24 120 720
```

```{r}
accumulate(1:6, `+`)

accumulate(1:6, `*`)
```

As for the rates example above, we concatenate the initial value with `catenate` 
(`,`) just like the R example, but otherwise this works fine

```apl
      rates ← 1.01 1.01 1.02 1.025 1.035 1.035 1.06
      inv ← 1000
      
      ×/inv, rates
1211.025672

      ×\inv, rates
1000 1010 1020.1 1040.502 1066.51455 1103.842559 1142.477049 1211.025672
```

So all of that recursive R code I wrote to generalise the cumulative application
of a function provided as an argument is boiled down to just the single glyph
`\`. Outstanding!

What's more, there are *lots* of binary functions one would use this with, all
of which have spelled-out names in other languages

```apl
      +/ ⍝ sum (add)
      ×/ ⍝ prod (multiply)
      ∧/ ⍝ all (and)
      ∨/ ⍝ any (or)
      ⌈/ ⍝ maximum (max)
      ⌊/ ⍝ minimum (min)
```

In summary, it seems that looking across these languages, the available options
range from a single glyph for `scan` along with the bare binary operator, e.g.
`×/`; a `cumprod()` function which isn't well-generalised but works out of the
box; an `accumulate=TRUE` argument to `Reduce`; and then there's whatever mess
this is (once you've _installed_ these)

```python
>>> from itertools import accumulate
>>> from operator import mul
>>> list(accumulate(rates, mul, initial=initial_investment))
```

Where did we go so wrong?

For what it's worth, Julia has a `reduce` and an `accumulate` that behave very 
nicely; generalised for the binary function as an argument

```julia
julia> reduce(+, 1:6)
21

julia> reduce(*, 1:6)
720

julia> accumulate(+, 1:6)
6-element Vector{Int64}:
  1
  3
  6
 10
 15
 21

julia> accumulate(*, 1:6)
6-element Vector{Int64}:
   1
   2
   6
  24
 120
 720
```

This is extremely close to the APL approach, but with longer worded names for
the `reduce` and `scan` operators. It also defines the more convenient `sum`,
`prod`, `cumsum`, and `cumprod`; no shortage of ways to do this in Julia!

In Haskell, `foldl` and `scanl` are the (left-associative) version of `reduce` 
and `accumulate`, and passing an infix as an argument necessitates wrapping it 
in parentheses

```haskell
ghci> foldl (+) 0 [1..6]
21

ghci> scanl (+) 0 [1..6]
[0,1,3,6,10,15,21]

ghci> foldl (*) 1 [1..6]
720

ghci> scanl (*) 1 [1..6]
[1,1,2,6,24,120,720]
```

This requires an explicit starting value, unless one uses the specialised 
versions which use the first value as an initial value

```haskell
ghci> foldl1 (+) [1..6]
21

ghci> scanl1 (+) [1..6]
[1,3,6,10,15,21]

ghci> foldl1 (*) [1..6]
720

ghci> scanl1 (*) [1..6]
[1,2,6,24,120,720]
```

I started this post hoping to demonstrate how nice the APL syntax was for this, 
but the detour through generalising the R function was a lot of unexpected fun 
as well.

Comments, improvements, or your own solutions are most welcome. I can be found
on [Mastodon](https://fosstodon.org/@jonocarroll) or use the comments below.

*Addendums*

It should probably be noted that R _does_ have a function `scan` but it's for 
reading data into a vector - if you ever spot someone using it for that... run. 
I have war stories about that function.

I'd love to hear how this is accomplished in some other languages, too - does it
have a built-in `accumulate` that takes a binary function?

Many thanks to commenters on Mastodon for reminding me that `Reduce(f, x,
accumulate=TRUE)` is the base R way to achieve the `scan`. I feel that still
counts as points against R because of the poor discoverability of gating the
_opposite_ behaviour behind a boolean flag.

We also have

```{r}
purrr::accumulate(1:6, `*`)
```

which is probably even more ergnomic, but I feel that takes away from my
complaints about needing to do an import in python.

<br />
<details>
  <summary>
    <tt>devtools::session_info()</tt>
  </summary>
```{r sessionInfo, echo = FALSE}
devtools::session_info()
```
</details>
<br />
