---
title: Digits Dilemma
author: Jonathan Carroll
date: '2024-06-15'
slug: digits-dilemma
categories:
  - rstats
  - julia
  - haskell
tags:
  - rstats
  - julia
  - haskell
type: ''
subtitle: ''
image: ''
editor_options: 
  chunk_output_type: console
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
  class.output  = "bg-success",
  class.message = "bg-info text-info",
  class.warning = "bg-warning text-warning",
  class.error   = "bg-danger text-danger"
)
```

Another day, another short riddle to be solved with several programming
languages! This one is nice because solving it doesn't need a lot of code, but 
it uses some interesting aspects of evaluation.

<!--more-->

Another day, another short riddle to be solved with several programming
languages! This one is nice because solving it doesn't need a lot of code, but 
it uses some interesting aspects of evaluation.

I saw [this post](https://jezenthomas.com/2022/05/solving-a-maths-riddle-with-bad-haskell/) 
which isn't new (it's from 2022) that poses a nice problem to solve:

> With the numbers 123456789, make them add up to 100. 
They must stay in the same order but you can use addition, subtraction, 
multiplication, division, brackets etc. All numbers must be used exactly once.

and demonstrates a solution in Haskell

```haskell
import Control.Monad (forM)
import Language.Haskell.Interpreter

expressions :: [String]
expressions =
  let ops = [ '+', '-', '/', '*' ]
   in [ [ '1', a, '2', b, '3', c, '4', d, '5', e, '6', f, '7', g, '8', h, '9' ]
        | a <- ops, b <- ops
        , c <- ops, d <- ops
        , e <- ops, f <- ops
        , g <- ops, h <- ops
        ]

result = runInterpreter $ do
  setImports ["Prelude"]
  exprs <- forM expressions evaluate
  pure $ filter (\(_, a) -> a == "100") $ fromRight [] exprs
  where
```

I'm trying to learn Haskell this year, so it was a great opportunity try to
follow along. I'm still working my way towards being able to run short snippets
- the `ghci` tool for interactive use has a bit of a learning curve and doesn't
immediately let me use the imports (or I'm doing something wrong) so while I
think I can follow the steps as presented, I can't yet dig into them as
interactively as I'd like.

The general idea, though, is to use a comprehension to expand all combinations of 
the allowed operators (`+`, `-`, `/`, and `*`) between the values 1 to 9. I'm 
somewhat familiar with comprehensions and played with them in my post  
[Pythagorean Triples with Comprehensions](https://jcarroll.com.au/2023/08/13/pythagorean-triples-with-comprehensions/)
in several languages, including Haskell.

I wanted to see how I might go about this problem in R, and I knew I'd need to make 
some adjustments because R does not have comprehensions. 

One way to get all the combinations of operators between the values is to use
`expand.grid()` which generates all combinations of its inputs

```{r}
expand.grid(1:3, letters[1:3])
```

Defining the operators as strings, I can generate a `data.frame` of the values and 
all combinations of operators between them 

```{r}
ops <- c("*", "+", "-", "/")
combos <- expand.grid(1, ops, 2, ops, 3, ops, 4, ops, 5, ops, 6, ops, 7, ops, 8, ops, 9)
head(combos)
```

This generates a lot of combinations - with 4 possible operators in 8 possible 
positions there are $4^8$ = 65,536 combinations. 

Pasting these numbers and operators together into expressions

```{r}
exprs <- apply(combos, 1, \(x) paste0(x, collapse = ""))
head(exprs)
```

I get something I can evaluate as if I typed `1*2*3` into a console. I can get 
the results of evaluating those with

```{r}
results <- sapply(exprs, \(x) eval(parse(text = x)))
head(results)
```

Now, I just need to see which of those produces a value of 100. Because `sapply` 
produced a vector with the expression itself as the name, I can extract the names
of the results which are equal to 100

```{r}
answers <- names(which(results == 100))
answers
```

Any of those can easily be verified manually

```{r}
1*2*3*4+5+6+7*8+9
```

One thing I noticed here was that I have one more result than the Haskell post 
produces

```{r}
length(answers)
```

One of the answers stands out in that it contains a division, and sure enough this 
is the one that doesn't appear in the Haskell post. I'm not quite sure why - I 
think the operator precedence is the same between R and Haskell, at least in terms 
of these expressions

```{r}
3 / 2 + 1
```

```haskell
ghci> 3 / 2 + 1
2.5
```

But since I haven't yet been able to actually _run_ that Haskell code myself, I can't 
verify those solutions.

My R solution to this puzzle is then

```r
ops <- c("*", "+", "-", "/")
combos <- expand.grid(1, ops, 2, ops, 3, ops, 4, ops, 5, ops, 6, ops, 7, ops, 8, ops, 9)
exprs <- apply(combos, 1, \(x) paste0(x, collapse = ""))
results <- sapply(exprs, \(x) eval(parse(text = x)))
names(which(results == 100))
```

I did want to try a language which _does_ have comprehensions - let's try Julia!

Setting up the comprehension makes for a bit of a long line, but comes out okay

```{julia}
ops = ['+', '-', '*', '/']
exprs = ['1' * a * '2' * b * '3' * c * '4' * d * '5' * e * '6' * f * '7' * g * '8' * h * '9' 
         for a in ops, b in ops, c in ops, 
             d in ops, e in ops, f in ops, 
             g in ops, h in ops];
first(exprs, 10)
```

That produces an array with many dimensions

```{julia}
size(exprs)
```

so it needs to be flattened into a vector with `vec()`. From there, it's similar
to the R approach and I can use a `eval(Meta.parse())` pattern, keeping in mind
that one can 'broadcast' scalar operations to vector operations with the dot
(`.`) operator

```{julia}
results = eval.(Meta.parse.(vec(exprs)));
first(results, 10)
```

Finding the values equal to 100 is similar to the R approach

```{julia}
exprs[findall(results .== 100)]
```

and again we see the 15 answers including the one with a division, confirming the 
R result. 

This was a fun exploration - I don't think I would want to try to solve it _without_
code, but the code solutions were a great opportunity to use a few different 
languages.

I suspect there are a few different ways of solving this apart from this brute-force 
expanding of every combination, perhaps with a solver or something. If you have one, I'd 
love to see it. I thought `@coolbutuseless` had done something like this but the closest
I could find was [this post](https://coolbutuseless.github.io/2021/12/09/lex-parse-and-evaluate-a-simple-s-expression-with-flexo/)
which is slightly different.

If you have comments, suggestions, or improvements, as always, feel free to use
the comment section below, or hit me up on
[Mastodon](https://fosstodon.org/@jonocarroll).

<br />
<details>
  <summary>
    <tt>devtools::session_info()</tt>
  </summary>
```{r sessionInfo, echo = FALSE}
devtools::session_info()
```
</details>
<br />
