---
title: Four Filters for Functional (Programming) Friends
author: Jonathan Carroll
date: '2023-08-30'
categories:
  - rstats
tags:
  - rstats
slug: four-filters-for-functional-programming-friends
editor_options:
  chunk_output_type: console
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
  class.output  = "bg-success",
  class.message = "bg-info text-info",
  class.warning = "bg-warning text-warning",
  class.error   = "bg-danger text-danger"
)
```

I'm part of a local Functional Programming Meetup group which hosts talks, but also 
coordinates social meetings where we discuss all sorts of FP-related topics including 
Haskell and other languages. We've started running challenges where we all solve a 
given problem in a language of our choosing then discuss over drinks how they compare. 

<!--more-->

I'm part of a local Functional Programming Meetup group which hosts talks, but also 
coordinates social meetings where we discuss all sorts of FP-related topics including 
Haskell and other languages. We've started running challenges where we all solve a 
given problem in a language of our choosing then discuss over drinks how they compare. 

This month we went with an "easy" problem with a wrinkle - we would solve the 'strain' 
exercise from Exercism ([Haskell](https://exercism.org/tracks/haskell/exercises/strain), 
[Python](https://exercism.org/tracks/python/exercises/strain) - your access to these is 
likely conditional on you being enrolled in that language track) with an extension:

> The problem is trivial; the challenge is to solve it in 4 different ways using your language of choice. 

The problem itself is given as

> Implement the `keep` and `discard` operation on collections. Given a collection and a predicate on the collection's elements, `keep` returns a new collection containing those elements where the predicate is true, while `discard` returns a new collection containing those elements where the predicate is false.
>
> For example, given the collection of numbers:
>
> `1, 2, 3, 4, 5`
>
> And the predicate:
>
> "is the number even?"
>
> Then your keep operation should produce:
>
> `2, 4`
>
> While your discard operation should produce:
>
> `1, 3, 5`


but with a restriction:

> Keep your hands off that filter/reject/whatchamacallit functionality provided by your standard library! Solve this one yourself using other basic tools instead. 

I figured it's a good opportunity to write as I solve it, so here's my R solutions.

I'll define a test case so I can try out things as I go

```{r}
test_vec <- c(1, 2, 3, 4, 5)
test_vec
```

and the predicates related to 'even' and 'odd' as functions which return `TRUE` or `FALSE`

```{r}
is_even <- function(x) {
  x %% 2 == 0
}

is_even(7)
is_even(8)

is_odd <- function(x) {
  !is_even(x)
}

is_odd(7)
is_odd(8)
```

Firstly, the restriction doesn't seem to worry me because when I think of "filter" in R 
I immediately think of `dplyr::filter()` which works on `data.frame` (or `tibble`) 
objects, and (given the examples) we're aiming to work with vectors (the problem is stated 
the same in several languages, so "collection" is a generalisation).

What about `base::Filter()`? The help states 

> `Filter` extracts the elements of a vector for which a predicate (logical) function gives true. 

```{r}
Filter(is_even, test_vec)
Filter(is_odd, test_vec)
```

Yep, that works exactly as I hoped, but is also a built-in "filter" so I can't use it. 

When I think of `keep` and `discard` I do think of the `purrr` functions, and while these, too 
do exactly what I want

```{r}
purrr::keep(test_vec, is_even)
purrr::discard(test_vec, is_even)
```

They're in a library, so I'm going to say they don't count.

One of the things I like about the way R does subsetting (via the square-bracket `[`
which _is by itself_ a function, but requires a matching `]` to satisfy the parser) is that you 
can use a logical vector to subset another vector, 

```{r}
c(3, 5, 8, 12)[c(TRUE, FALSE, FALSE, TRUE)]
```

which means that if I can _produce_ such a logical vector, say, by applying a predicate function, I can do subsetting that way 

```{r}
test_vec[is_even(test_vec)]
test_vec[is_odd(test_vec)]
```

Instead of using `is_odd()` I can just _negate_ the logical vector to get the same effect

```{r}
test_vec[!is_even(test_vec)]
```

I can make those into functions that take a predicate and a vector

```{r}
keep_1 <- function(f, x) {
  x[f(x)]
}

discard_1 <- function(f, x) {
  x[!f(x)]
}
```

Testing these

```{r}
keep_1(is_even, test_vec)
discard_1(is_even, test_vec)
```

One down! 

One thing to note with this approach is that R is _vectorised_ - I've discussed this 
a few times on this blog ([most recently](https://jcarroll.com.au/2023/08/29/now-you-re-thinking-with-arrays/)) - which 
means that these predicate functions will gladly take a vector, not just a single value. This 
works for the `is_even()` function because inside that, the modulo operator `%%` is itself 
vectorised, so 

```{r}
is_even(c(2, 4, 6, 9, 11, 13))
```

As I wrote in my previous post, thinking like this just becomes so natural in R that I 
have to force myself to remember that not every language does that.

It's also worth mentioning that I'm passing a reference to the function `is_even` to our 
`keep` and `discard` functions - that is serving as our predicate because I need a way to 
state "is the number even?" which references the number, so I need a function. That _doesn't_ 
have to be a named function, though - it could be an "anonymous" function (a "lambda") if 
I wanted

```{r}
keep_1(function(z) z %% 2 == 0, test_vec)
discard_1(function(z) z %% 2 == 0, test_vec)
```

I can subset a vector with a logical vector of the same length, specifying whether or not 
to include that element, but I can _also_ subset by position (keeping in mind that R is a 
1-based language which means the first element is indexed by a `1` - why would any language 
do anything different? 😜)

```{r}
c(3, 5, 8, 12)[c(1, 4)]
```

The function `which()` takes a logical vector and returns which indices are `TRUE`

```{r}
which(c(TRUE, FALSE, FALSE, TRUE))
```

so I can use this with our predicate to keep elements

```{r}
keep_2 <- function(f, x) {
  x[which(f(x))]
}

keep_2(is_even, test_vec)
```

However, discarding elements by index doesn't use a logical negation, it uses a 
negative sign (`-`)

```{r}
discard_2 <- function(f, x) {
  x[-which(f(x))]
}

discard_2(is_even, test_vec)
```

If you look at the source of `Filter()`, you'll see that I wasn't far off of exactly 
that

```{r}
Filter
```

but it still counts. 

Another option would be to unpack the elements themselves and do some stepwise 
comparisons in a loop. For each element `el` in the vector `x`, test if `f(el)` is `TRUE`, 
and if it is, concatenate `el` to the end of the accumulating `result` vector

```{r}
keep_3 <- function(f, x) {
  result <- c()
  for (el in x) {
    if (f(el)) {
      result <- c(result, el)
    }
  }
  result
}

keep_3(is_even, test_vec)

discard_3 <- function(f, x) {
  result <- c()
  for (el in x) {
    if (!f(el)) {
      result <- c(result, el)
    }
  }
  result
}

discard_3(is_even, test_vec)
```

Of course, this approach is a Bad Idea™ in general but I'm not optimising anything here. This 
approach _does_ have the advantage that it isn't relying on R's vectorised capabilities, since 
each element is passed to the predicate function individually, so if I did have a non-vectorized 
predicate function, this would still work.

I really want a "weird" way to do this. R has plenty of weird to go around, but since I've 
been learning some Haskell, and the challenge originally referenced the Haskell solution, 
what if I code a Haskell-esque solution? 

Haskell makes good use of recursive functions. Any loop can be written as a recursion 
(and vice-versa) so the previous solution is a good starting point. First, I define 
a base case; if I run out of numbers to process, return `NULL`. A convenient feature 
of R vectors is that `NULL`s are dropped

```{r}
c(1, 2, NULL, 3, 4, NULL, 5)
```

Otherwise, I can take the first value in the vector and test it with the predicate. If it 
returns `TRUE` I can append it to what I've calculated so far, and recursively 
call the function again with the rest of the vector. That could look like

```{r}
keep_4 <- function(f, x) {
  if (!length(x)) return(NULL)
  if (f(x[1])) {
    return(c(x[1], Recall(f, x[-1])))
  } else {
    return(Recall(f, x[-1]))
  }
}

keep_4(is_even, test_vec)
```

Some interesting points about this: the `Recall()` function is nice for defining a recursive 
function. I could have used `keep_4` there, but the advantage of this implementation is 
that I can rename the function and it still works as expected

```{r}
keep_4_also <- keep_4
rm("keep_4")

keep_4_also(is_even, test_vec)
```

If I had explicitly referenced `keep_4` inside itself, that recursion would fail with this 
renaming. 

The negative subsetting works as described above; `x[-1]` means "not including the first
element". Lastly, testing `if (!length(x))` works because `0` can be coerced to `FALSE` and 
any other value to `TRUE`, so if the `length` of `x` is _not_ `0`, this condition is met.

The discarding variant is similar, just with the two `returns()` around the other way, or

```{r}
discard_4 <- function(f, x) {
  if (!length(x)) return(NULL)
  if (!f(x[1])) {
    return(c(x[1], Recall(f, x[-1])))
  } else {
    return(Recall(f, x[-1]))
  }
}

discard_4(is_even, test_vec)
```

There we go; 4 hand-coded implementations of `keep` and `discard` in R.

Can you think of another that doesn't use `Filter()` or an external library? Let me 
know in the comments below or on [Mastodon](https://fosstodon.org/@jonocarroll). I'm 
looking forward to seeing how people solved this in other languages.

<br />
<details>
  <summary>
    <tt>devtools::session_info()</tt>
  </summary>
```{r sessionInfo, echo = FALSE}
devtools::session_info()
```
</details>
<br />
