---
title: 'Project Euler Q3 :: Largest prime factor'
author: Jonathan Carroll
date: 2015-01-03 00:52:07
slug: project-euler-q3
categories: [R]
tags: [rstats]
type: ''
subtitle: ''
image: ''
editor_options: 
  chunk_output_type: console
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
class.output= "bg-success",
class.message = "bg-info text-info",
class.warning = "bg-warning text-warning",
class.error= "bg-danger text-danger"
)
```

[Explanation](https://jcarroll.com.au/2015/01/02/project-euler/). Standard caveat: don't look here if you are trying to do these yourself.

<!--more-->

[Explanation](https://jcarroll.com.au/2015/01/02/project-euler/). Standard caveat: don't look here if you are trying to do these yourself.

<blockquote>
The prime factors of 13195 are 5, 7, 13 and 29.

What is the largest prime factor of the number 600851475143?
</blockquote>

It seems so simple at first glance, until of course you look at how big that last number is. I started off by making sure I understood the issue.

```{r}
## check the worked solution
5*7*13*29
```

so far so good. At this point I realised that there isn't an inbuilt `is.prime` so I stole one from [this site.](http://cran.r-project.org/web/packages/rlist/vignettes/Examples.html)

```{r}
is.prime <- function(num) {
 if (num == 2L) {
 TRUE
 } else if (any(num %% 2L:(num-1L) == 0L)) {
 FALSE
 } else {
 TRUE
 }
}
```

Testing the example works pretty well...

```{r}
## let's loop up to n and list the prime factors
prime.factors <- function(n) {
 primes <- c()
 for(i in 1:n) {
 ## take advantage of lazy logical evaluation
 ## and short-cut to only the factors
 if(n %% i == 0 & is.prime(i)) primes <- c(primes, i)
 }
 return(primes)
}
prime.factors(13195)
```

but I hit a snag when I tried to do the same for the problem value.

```{r, error = TRUE}
w <- as.integer(600851475143)
prime.factors(600851475143) ## Error: cannot allocate vector of size 4476.7 Gb
```

Sure enough, that's bigger than the machine precision integer allows

```{r}
as.numeric("600851475143") > .Machine$integer.max
```

so, I abandoned the pre-filled list of values and went again with the brute force. For the sake of speeding it up, I delayed testing for primes until later, as I can do that over the generated list with an `apply` and only bothered testing the values below `sqrt(n)` and `n/f` where `f` is the largest found prime so far.

```{r}
## lists are too big. Find the primes by brute force
## using floating point representations
z <- as.numeric("600851475143")
i <- 2
factors <- 1
## loop through values of i that are
## less than sqrt(z) and
## less than z/the largest found factor
while(i < sqrt(z) & i < z/max(factors)) {
 ## skip the prime test for now
 if(z %% i == 0) factors <- c(factors, i)
 i <- i + 1
}
factors
factors.prime <- sapply(factors, is.prime)
primes <- factors[factors.prime] 
z == prod(primes)

max(primes)

### CORRECT
```

<br />
<details>
  <summary>
    <tt>devtools::session_info()</tt>
  </summary>
```{r sessionInfo, echo = FALSE}
devtools::session_info()
```
</details>
<br />

