---
title: 'Project Euler Q1 :: Multiples of 3 and 5'
author: Jonathan Carroll
date: 2015-01-02 22:45:10
slug: project-euler-q1
categories: [R]
tags: [rstats]
type: ''
subtitle: ''
image: ''
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
class.output= "bg-success",
class.message = "bg-info text-info",
class.warning = "bg-warning text-warning",
class.error= "bg-danger text-danger"
)
```

[Explanation](http://jcarroll.com.au/2015/01/code/project-euler/). Standard caveat: don't look here if you are trying to do these yourself.

<!--more-->

[Explanation](http://jcarroll.com.au/2015/01/code/project-euler/). Standard caveat: don't look here if you are trying to do these yourself.

<blockquote>
If we list all the natural numbers below 10 that are multiples of 3 or 5, we get 3, 5, 6 and 9. The sum of these multiples is 23. Find the sum of all the multiples of 3 or 5 below 1000.
</blockquote>

This one's pretty straight forward, really, as one might hope being the first question. R's built-in subsetting mechanism handles the extraction fairly nicely. I perhaps would have liked a way to do this without first defining <code>x</code>; though I suppose it could just be repeated in the last line.

```{r}
## check the worked solution
sum(c(3,5,6,9))

## values < 1000
x <- 1:999

## sum of x % 3 or x % 5
sum(x[x %% 3 == 0 | x %% 5 == 0])

### CORRECT
```

<br />
<details>
  <summary>
    <tt>devtools::session_info()</tt>
  </summary>
```{r sessionInfo, echo = FALSE}
devtools::session_info()
```
</details>
<br />
