---
title: 'Project Euler Q4 :: Largest palindrome product'
author: Jonathan Carroll
date: 2015-01-08 22:24:09
slug: project-euler-q4
categories: [R]
tags: [rstats]
type: ''
subtitle: ''
image: ''
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
class.output= "bg-success",
class.message = "bg-info text-info",
class.warning = "bg-warning text-warning",
class.error= "bg-danger text-danger"
)
```

[Explanation](https://jcarroll.com.au/2015/01/02/project-euler/). Standard caveat: don't look here if you are trying to do these yourself.

<!--more-->

[Explanation](https://jcarroll.com.au/2015/01/02/project-euler/). Standard caveat: don't look here if you are trying to do these yourself.

<blockquote>
A palindromic number reads the same both ways. The largest palindrome made from the product of two 2-digit numbers is 9009 = 91 × 99. Find the largest palindrome made from the product of two 3-digit numbers.
</blockquote>

This seems like another brute-force question. There's not that many numbers to test.

```{r}
## check the worked solution
91*99 
```

I'm not aware of an `is.palindrome` function, but it's easy enough to code.

```{r}
is.palindrome <- function(x) {
   ## convert to character and explode
   x <- unlist(strsplit(as.character(x), ""))
   ## check if the vector is palindromic
   return(identical(x, rev(x)))
}
is.palindrome(9009)
is.palindrome(9001)
```

Let's try it out for the two digit example and make sure we're on the right track. Multiply all two digit numbers together and test them for palindrome-ness, then find the largest of those.

```{r}
twodigits <- 10:99
prods <- expand.grid(twodigits, twodigits)
prods$prod <- prods[ ,1]*prods[ ,2]
prods.palindromes <- prods$prod[sapply(prods$prod, is.palindrome)]
max(prods.palindromes)
```

Great! What about three digits?

```{r}
threedigits <- 100:999
prods <- expand.grid(threedigits, threedigits)
prods$prod <- prods[ ,1]*prods[ ,2]
prods.palindromes <- prods$prod[sapply(prods$prod, is.palindrome)]
largest <- max(prods.palindromes)
largest

### CORRECT
```

Takes a little longer, and generates a nice little 10MB, 810,000 element vector along the way.

```{r}
format(object.size(prods), units="Mb")
```

The two three digit numbers?

```{r}
prods[prods$prod==largest, ]
```

<br />
<details>
  <summary>
    <tt>devtools::session_info()</tt>
  </summary>
```{r sessionInfo, echo = FALSE}
devtools::session_info()
```
</details>
<br />

