---
title: 'Project Euler Q5 :: Smallest multiple'
author: Jonathan Carroll
date: 2015-01-08 23:25:58
slug: project-euler-q5
categories: [R]
tags: [rstats]
type: ''
subtitle: ''
image: ''
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
class.output= "bg-success",
class.message = "bg-info text-info",
class.warning = "bg-warning text-warning",
class.error= "bg-danger text-danger"
)
```

[Explanation](http://jcarroll.com.au/2015/01/code/project-euler/). Standard caveat: don't look here if you are trying to do these yourself.

<!--more-->

[Explanation](http://jcarroll.com.au/2015/01/code/project-euler/). Standard caveat: don't look here if you are trying to do these yourself.

<blockquote>
2520 is the smallest number that can be divided by each of the numbers from 1 to 10 without any remainder. What is the smallest positive number that is _evenly divisible_ by all of the numbers from 1 to 20?
</blockquote>

I'm getting the feeling that brute-force is going to be quite the useful tool for these questions. Thankfully `R` can churn through numbers really fast.

So, we're after a number divisible by `1, 2, 3, ..., 10`. Let's vectorise that and check the stated answer

```{r}
all(2520 %% 1:10 == 0)
```

Easy enough. The solution value must be divisible by 20, so we can just test multiples of 20 for the above property

```{r}
i <- 20
y <- FALSE
while(!y) {
  i <- i + 20
  y <- all(i %% 1:20 == 0)
}
i

### CORRECT
```

Wrapping a `system.time()` call around that assures us that this is still done in under a minute, as per the guidelines

```r
   user  system elapsed 
 26.150   0.000  26.192 
```

<br />
<details>
  <summary>
    <tt>devtools::session_info()</tt>
  </summary>
```{r sessionInfo, echo = FALSE}
devtools::session_info()
```
</details>
<br />

