---
title: Print Debugging (Now with Icecream!)
author: Jonathan Carroll
date: '2023-11-07'
slug: print-debugging-now-with-icecream
categories:
  - rstats
  - rust
tags:
  - rstats
  - rust
type: ''
subtitle: ''
image: ''
editor_options: 
  chunk_output_type: console
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
  class.output  = "bg-success",
  class.message = "bg-info text-info",
  class.warning = "bg-warning text-warning",
  class.error   = "bg-danger text-danger"
)
```

Print debugging has its place. Sure, it's not always the *best* way to debug something, 
but it can often be the fastest. In this post I describe a useful way to do this in Rust 
and how we can get similar behaviour in R.

<!--more-->

Print debugging has its place. Sure, it's not always the *best* way to debug something, 
but it can often be the fastest. In this post I describe a useful way to do this in Rust 
and how we can get similar behaviour in R.

I love the Rust [`dbg!`](https://dev-doc.rust-lang.org/beta/std/macro.dbg.html)
macro - it wraps a value or expression and prints the result to help debug
what's happening in the middle of some function. If we had some complicated
function that combined some values, e.g.

```rust
fn f(val1: i32, val2: i32) -> i32 {
    // do some things
    let otherval: i32 = 10;
    // final result
    val1 + val2 + otherval
}

fn main() {
    println!("final result = {}", f(5, 6))
}
```

Running this gives just the final result, as expected

```
final result = 21
```

We might want to check on what the intermediate combinations of `otherval` and `val1` 
or `val2` (terrible names, I know). One option is to just print them

```rust
fn f(val1: i32, val2: i32) -> i32 {
    // do some things
    let otherval: i32 = 10;
    println!("{}", otherval + val1);
    println!("{}", otherval + val2);
    // final result
    val1 + val2 + otherval
}

fn main() {
    println!("final result = {}", f(5, 6))
}
```

Running this shows the values we printed, but with no context

```
15
16
final result = 21
```

We could add some context manually

```rust
fn f(val1: i32, val2: i32) -> i32 {
    // do some things
    let otherval: i32 = 10;
    println!("first temp val = {}", otherval + val1);
    println!("second temp val = {}", otherval + val2);
    // final result
    val1 + val2 + otherval
}

fn main() {
    println!("final result = {}", f(5, 6))
}
```

producing

```
first temp val = 15
second temp val = 16
final result = 21
```

but across an entire codebase, this is going to get messy, fast.

A better option is the `dbg!` macro, which takes an _expression_ (a value, or 
something that produces a value) and prints both the expression itself and the resulting 
value

```rust
fn f(val1: i32, val2: i32) -> i32 {
    // do some things
    let otherval: i32 = 10;
    dbg!(otherval + val1);
    dbg!(otherval + val2);
    // final result
    val1 + val2 + otherval
}

fn main() {
    println!("final result = {}", f(5, 6))
}
```

Running this produces

```
[src/main.rs:15] otherval + val1 = 15
[src/main.rs:16] otherval + val2 = 16
final result = 21
```

and we see we get the file/linenumber context, the expression we wrapped, and the value.

This is extremely useful, and helps me to figure out what's going on in the middle of 
some implementation.

One of the downsides is that even if I make a release build, these statements remain, 
so I need to go through and remove them all once I'm finished debugging.

A "better" solution is to use a full logging solution like the
[`log`](https://docs.rs/log/latest/log/) crate which enables using different log
levels, turning off logging outside of some threshold, etc... but that seems
more suited to intentional logging, not debugging something that _isn't_
working.

Having played with this in Rust, of course I wanted the same thing in R. I built a 
minimal viable proof-of-concept which leverages {rlang} to capture the expression

```{r}
dbg <- function(x) {
  ex <- rlang::f_text(rlang::enquos(x)[[1]])
  ret <- rlang::eval_bare(x)
  message(glue::glue("DEBUG: {ex} = {ret}"))
  ret
}
```

This works rather well - it postpones evaluation of the expression and prints the result 
without affecting any variables

```{r}
a <- 1
b <- 3
x <- dbg(a + b)
y <- dbg(2*x + 3)
z <- 10 + dbg(y*2)
z
```

It lacks one beautiful part of the Rust solution, though - if I include this in some 
functions sourced from a file, I won't be able to tell which file the statement came from. Plus, 
it doesn't deal so well with large structures

```{r}
x <- dbg(head(mtcars))
```

At some point I saw [a blog
post](https://turtletopia.github.io/2022/07/28/ice-cream-for-r-programmers/)
about a debug logging package [{icecream}](https://github.com/lewinfox/icecream)
which had this ability of tracing the `srcref` of a file containing the debug
statement, so I wanted to see if I could extract that to suit my needs. Running
the `ic()` statement as a print debugger works nicely

```{r, include=FALSE}
options(
    icecream.enabled = TRUE,
    icecream.prefix = "ic|",
    icecream.output.function = NULL,
    icecream.peeking.function = icecream:::ic_autopeek,
    icecream.max.lines = 1,
    icecream.arg.to.string.function = NULL,
    icecream.always.include.context = FALSE
  )
```

```{r}
f <- function(val1, val2) {
  otherval <- 10
  icecream::ic(otherval + val1)
  icecream::ic(otherval + val2)
  val1 + val2 + otherval
}
f(5, 6)
```

it indeed wraps the expression and shows the result. After poking around at the 
internals, I realised it actually does everything that I wanted, I just needed to change 
some of the defaults

```{r}
options(icecream.peeking.function = utils::head,
        icecream.max.lines = 5,
        icecream.prefix = "dbg:",
        icecream.always.include.context = TRUE)
```

Now it prints like the `dbg!` macro

```{r}
f <- function(val1, val2) {
  otherval <- 10
  icecream::ic(otherval + val1)
  icecream::ic(otherval + val2)
  val1 + val2 + otherval
}
f(5, 6)
```

To make it even more like the Rust macro, I made a similar binding of `.dbg` (so
that it doesn't appear in my workspace by default) and added the following to my
.Rprofile

```
# install.packages("icecream")
if (requireNamespace("icecream", quietly = TRUE)) {
  .dbg <- icecream::ic
  options(icecream.peeking.function = utils::head,
          icecream.max.lines = 5,
          icecream.prefix = "dbg:",
          icecream.always.include.context = TRUE)
}
```

so now I can add debug statements

```{r}
f <- function(val1, val2) {
  otherval <- 10
  .dbg(otherval + val1)
  .dbg(otherval + val2)
  val1 + val2 + otherval
}
f(5, 6)
```

Better yet, I can turn them off if I don't need them

```{r}
icecream::ic_disable()
f(5, 6)
```

This works as I had hoped; I can even debug partway through an expression

```{r}
icecream::ic_enable()
3 + .dbg(4 + 6)
```

and if I source a file, I get the context

```
## test_dbg.R:
f <- function() {
  x <- 7
  .dbg(x + 3)
  7
}
```

```{r, eval = FALSE}
source("test_dbg.R")
f()
```

```{r, eval = FALSE, class.source = "bg-info"}
ℹ dbg: `f()` in test_dbg.R:3:2 | `x + 3`: [1] 10
```
```{r, eval = FALSE, class.source = "bg-success"}
[1] 7
```

It even deals with printing larger objects, given the "peeking_function" and "max lines" 
options above

```{r}
.dbg(mtcars)
```

That seems to be feature-equivalent to the Rust `dbg!` macro, plus the ability to 
turn off the messages, so I'm very happy with this result.

Do you have a better solution? I can be found on
[Mastodon](https://fosstodon.org/@jonocarroll) or use the comments below.

<br />
<details>
  <summary>
    <tt>devtools::session_info()</tt>
  </summary>
```{r sessionInfo, echo = FALSE}
devtools::session_info()
```
</details>
<br />
