---
title: In-Place Modifications
author: Jonathan Carroll
date: '2024-09-25'
categories:
  - APL
  - python
  - rstats
  - julia
tags:
  - APL
  - python
  - rstats
  - julia
slug: in-place-modifications
editor_options:
  chunk_output_type: console
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
  class.output  = "bg-success",
  class.message = "bg-info text-info",
  class.warning = "bg-warning text-warning",
  class.error   = "bg-danger text-danger"
)
```

In this post I explore some differences between R, python, julia, and APL in 
terms of mutability, and try to make something that probably shouldn't exist.

<!--more-->

In this post I explore some differences between R, python, julia, and APL in 
terms of mutability, and try to make something that probably shouldn't exist.

I watched [this code_report](https://youtu.be/O9AsxoQzNdI) video which describes 
a [leetcode problem](https://leetcode.com/problems/final-array-state-after-k-multiplication-operations-i/description/); 

> You are given an integer array `nums`, an integer `k`, and an integer `multiplier`.
> 
> You need to perform `k` operations on `nums`. In each operation:
> 
> * Find the minimum value `x` in `nums`. If there are multiple occurrences of the minimum value, select the one that appears first.
> * Replace the selected minimum value `x` with `x * multiplier`.
>
> Return an integer array denoting the final state of nums after performing all `k` operations.

Conor's python solution in the video was

```{python}
def getFinalState(nums, k, m): 
  for _ in range(k): 
    i = nums.index(min(nums)) 
    nums[i] *= m
  return nums

x = [2, 1, 3, 5, 6]
k = 5
mult = 2

getFinalState(x, k, mult)
```

and, as always, I wanted to see how I'd do that in R. I came up with this

```{r}
getFinalState = function(nums, k, m) {
  for (i in 1:k) {
    m <- which.min(nums)[1]
    nums[m] <- mult * nums[m]
  }
  nums
}

x <- c(2, 1, 3, 5, 6)
k <- 5
mult <- 2

getFinalState(x, k, mult)
```

It's worth noting that I can't use a `map` in this function because iterations
are dependent; the minimum value at any iteration depends on the previous
values.

I also had a chance to discuss this solution with some APL'ers at a meetup and 
a J solution was presented, but I don't think I wrote it down.

My solution is nearly word-for-word the same as the python solution with a
couple of notable exceptions arising from the difference between the two
languages:

First, R has `which.min()` as a built-in rather than needing to query the index
of the minimum value (and two references to `nums`). Also, R has no compound
assignment like `x *= 2` which modifies in-place - the closest thing I can think
of is the `%<>%` operator in {magrittr} (not re-exported in {dplyr} because this
behaviour is considered bad practice in R, despite not really being "in-place")

```{r}
library(magrittr)

m <- data.frame(x = 1:6, y = letters[1:6])
m

m %<>% head(2)
m
```

although I can certainly see the case for it - this operator avoids repeating
the variable being used and assigned, because the alternative using the
traditional pipe is

```{r}
m <- data.frame(x = 1:6, y = letters[1:6])
m

m <- m %>% head(2)
m
```

One could argue that writing out even a longer variable name twice still makes 
it clear that shadowing is taking place; the value is being overwritten with 
a new value, but it does feel a little frustrating to have to type it out twice

```r
important_variable <- important_variable * 2
```

Back to my R solution, the indexing at a specific set of values got me thinking
that it would be clean if we could pass a _function_ to `[` so that we could 
write

```r
nums[which.min] <- value
```

(maybe not so much for this example where `m` is used twice, but it piqued my
interest)

Let's say I want to set all the even values of a vector to some other value.
That's easy enough to do

```r
x[x %% 2 == 0] <- 0
```

but I don't love that it requires two references to `x`, which may (should?) be
a much longer name

```r
important_variable[important_variable %% 2 == 0] <- 0
```

I want something like `x[f] <- y` to set the values of `x` where `f(x)` is
`TRUE` to `y`. This seemed like it might be possible, maybe with a `function`
method to `[<-`, but `[<-` dispatches on the class of `x`, not what's inside
`[`, so no dice. In theory (which will _never_ happen) the built-in `[<-` could
have some branch logic for dealing with a function passed as the indices to be
modified, but I'm not about to go rebuilding R from source myself just to play
with that idea.

Nonetheless, if I define some functions that do accomplish this

```{r}
is_even <- function(z) z %% 2 == 0

set_if <- function(x, f, value) {
  x[f(x)] <- value
  x
}
```

then I can try this out on a vector

```{r}
a <- 1:10
a
set_if(a, is_even, 0)
a # unchanged
```

It works, but I'm back to having to write `a <- do_stuff(a)` because `a` isn't 
actually modified by this function.

Ideally, my function would operate the same as this does

```{r}
a <- 1:10
a[is_even(a)] <- 0
a
```

which _does_ modify `a` in-place; R is not entirely pure, and does occasionally 
allow what looks like direct mutation, though under the hood, it's not - a new 
object is actually created

```{r}
# not using a range e.g. 1:n because that's internally 
# a "compact" representation
a <- c(2, 3, 4)
.Internal(inspect(a))

a[2] <- 9
.Internal(inspect(a))
```

Note that the memory address has changed. 

If I was working with a language which did support (enable?) modify-in-place
then that might look like

```{python}
def is_even(x):
   return x % 2 == 0

def set_if(x, f, value):
     for i in range(len(x)):
         if f(x[i]):
             x[i] = value

a = list(range(10))
a

set_if(a, is_even, 0)
a
```

Now, that's not always a great thing. In such a language with mutable structures
(e.g. lists) we can do maddening things like this

```{python}
x = [3, 4, 5]
y = x
y is x
y[1] = 9
x # still 'bound' to y
```

Here, `is` means "are these two things identical in the sense of referring to 
the same block of memory", noting that literals (e.g. single numbers) are 
referenced that way, but tuples aren't

```{python}
abc = (11, 99)
xyz = (11, 99)
abc is xyz
abc == xyz
```

The big question is can I hack together some solution that _does_ work in-place
in R? Yeah, with some ill-advised calls

```{r}
set_if <- function(x, f, value) {
  # can't use <<- because the value passed in as the x argument isn't 
  # necessarily named 'x' in the parent scope
  .x <- x
  .x[f(.x)] <- value
  e <- parent.env(environment())
  assign(deparse(substitute(x)), .x, pos = e)
  invisible(.x)
}


a <- 1:10
a
set_if(a, is_even, 0)
a
```

As I note in the comment there, I can't use the super-assignment arrow `<<-` 
inside this function because I don't know the name of the variable I'm updating; 
it needs to be deparsed from the incoming argument.

This means that it works regardless of the name of the variable being modified

```{r}
b <- 10:20
b
set_if(b, is_even, 0)
b
```

I tried to think of some other languages which might support this sort of in-place 
`set_if(x, f, value)` modification and (Dyalog) APL was worth a thought.

```apl
    ⍝ create a vector from 1 to 10
    x←⍳10
    x
1 2 3 4 5 6 7 8 9 10

    ⍝ the function {0=2|⍵} calculates a boolean vector with 
    ⍝ 1 where the value is even
    {0=2|⍵} x
0 1 0 1 0 1 0 1 0 1

    ⍝ the `@` operator takes a value (or function) on the left and 
    ⍝ a function (or boolean values) on the right and applies it to the 
    ⍝ other argument on the right
    0@{0=2|⍵} x 
1 0 3 0 5 0 7 0 9 0

    ⍝ alternatively a point-free function defined as the negation (`~`) of a 
    ⍝ binding (`∘`) of the value 2 to modulo (`|`); the negation is needed
    ⍝ otherwise this returns the result of the modulo, not where it is 0
    0@(~2∘|)⍳10
1 0 3 0 5 0 7 0 9 0

    ⍝ x is, however, unchanged as APL is typically immutable
    x
1 2 3 4 5 6 7 8 9 10
```

So there's no way to do the in-place modification. it is nice, though, that
`0@(~2∘|)x` only refers to `x` once.

Julia makes a nice distinction between functions which mutate arguments and 
those which don't; (by convention) the former are named ending with an 
exclamation mark, e.g.

```{julia}
vec = collect(1:5)
# non-mutating
reverse(vec)
vec
# mutating
reverse!(vec)
vec
```

In julia, the `iseven()` function is already built-in, but vectorisation is 
explicit via a broadcast operator `.` and the setting of even values to 0 
looks like

```{julia}
x = collect(1:10);
x[iseven.(x)] .= 0;
x
```

which looks very much like the R version with some dots where scalar functions 
are vectorised. If I don't use the last `.` to perform vectorised assignment, 
the error tells me that the failure involved the `setindex!` function which does 
sound like what I want, but this doesn't work

```julia
setindex!(x, 0, iseven.(x))
```

because it's trying to assign the value 0 multiple times and I only provided one 
of them. Instead,

```{julia}
x = collect(1:10);
setindex!(x, zeros(Int8, 5), iseven.(x));
x
```

does work, but I had to manually count how many 0 entries this requires, so the
`[` approach seems cleaner. Either way, I've had to explicitly calculate
`iseven(x)` and pass that result somewhere.

Since Julia allows users to extend methods, I _could_ do that modification myself!

```{julia}
import Base.setindex! 
  
function setindex!(A::Vector{Int64}, v::Int64, f::Function) 
  A[f.(A)] .= v
end
x = collect(1:10);
setindex!(x, 0, iseven);
x
```

which I could just as easily call `set_if!` 

```{julia}
set_if! = setindex!;
x = collect(1:10);
set_if!(x, 0, iseven);
x
```

Nice! I do wonder if I can "hack" (ahem, extend) Julia's `[` to get my prized
`x[f] = 0` solution but I doubt it's worth it when the above does the right
thing.

I don't imagine I'll package up my `set_if()` anywhere, and I should probably 
even avoid using it myself, but it's been an interesting journey thinking about 
this stuff. Maybe there's a better way to do it? Maybe there's a language which 
better supports something like that? If you know, or you have comments or 
suggestions, I can be found on
[Mastodon](https://fosstodon.org/@jonocarroll) or use the comment section below.

**UPDATE** After some inspiration from Doug Kelkhoff I did end up implementing 
this as a real package: [{vec}](https://github.com/jonocarroll/vec) is intended 
as a new vector class where we could implement new features. There are already 
a couple of suggestions, and if you have something you wish R's vectors supported, 
then please add an Issue!

**UPDATE2** I probably should have noted that the original motivation of updating 
the minimum value from the code_report video _does_ indeed work with this new 
package

```{r}
library(vec)

nums <- as_vec(c(4, 5, 3, 6, 8))
nums[which.min] <- 99
nums
```

and I'm very happy about that!

<br />
<details>
  <summary>
    <tt>devtools::session_info()</tt>
  </summary>
```{r sessionInfo, echo = FALSE}
devtools::session_info()
```
</details>
<br />
