---
title: Taking from Infinite Sequences
author: Jonathan Carroll
date: '2023-08-18'
slug: taking-from-infinite-sequences
categories:
  - rstats
  - haskell
  - python
  - rust
tags:
  - rstats
  - haskell
  - python
  - rust
type: ''
subtitle: ''
image: ''
editor_options: 
  chunk_output_type: console
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
  class.output  = "bg-success",
  class.message = "bg-info text-info",
  class.warning = "bg-warning text-warning",
  class.error   = "bg-danger text-danger"
)
```

One thing that has really caught my attention as I learn more programming 
languages is the idea of _generators_ or infinite sequences of values. Yes, _infinite_.
Coming from R, that seems unlikely, but in at least several other languages, 
it's entirely possible thanks to iterators and lazy evaluation.

<!--more-->

One thing that has really caught my attention as I learn more programming 
languages is the idea of _generators_ or infinite sequences of values. Yes, _infinite_.
Coming from R, that seems unlikely, but in at least several other languages, 
it's entirely possible thanks to iterators and lazy evaluation.

First, a diversion into recursion

![A timely combination. [\@rverbsr](https://fosstodon.org/@rverbsr@botsin.space/110903654592091528)](images/recursive.png)

In Haskell, one of the first exercises after "Hello, World!" people discover is 
the Fibonacci sequence, where the $n^{\rm th}$ value is given by the sum of the 
two previous values. As a function, this can be written as

```haskell
λ> fib 0 = 0
λ> fib 1 = 1
λ> fib n = fib (n-1) + fib (n-2)
```

Essentially, `fib(0)` returns `0`. `fib(1)` returns `1`, and for any value `n` it 
returns the (recursively defined) sum of the two previous values. This isn't infinite
at all...

$$
\begin{align*}
{\rm fib}(4) &= {\rm fib}(3) + {\rm fib}(2)\\
&= ({\rm fib}(2) + {\rm fib}(1)) + ({\rm fib}(1) + {\rm fib}(0))\\
&= {\rm fib}(1) + {\rm fib}(0) + {\rm fib}(1) +  {\rm fib}(1) + {\rm fib}(0)\\
&= 1 + 0 + 1 + 1 + 0\\
&= 3
\end{align*}
$$

We could write that in R as

```{r}
fib <- function(n) {
  if (n == 0) return(0)
  if (n == 1) return(1)
  fib(n - 1) + fib(n - 2)
}

fib(4)

fib(8)
```

This may come as a surprise to some - the function is defined _in terms of itself_ and 
some base cases. This is entirely fine in R and Haskell as they're lazily evaluated - 
nothing happens until a value is actually used. In R, this means that if an argument 
to a function isn't used, it's not evaluated at all

```{r}
loud_func <- function() {
  cat("HELLLOOOOO!!!\n")
}

stays_quiet <- function(f, g = loud_func()) {
  f(c(1, 2, 3, 4))
}

stays_quiet(f = mean)
```

Notice that although the argument `g` is the _invocation_ of `loud_func()`, it's never
evaluated because we don't use it. If we did use it...

```{r}
noisy <- function(f, g = loud_func()) {
  g
  f(c(1, 2, 3, 4))
}

noisy(sum)
```

So, in these languages, we can define a function recursively. Given the base case(s), 
these will eventually return just a number, so the computation will complete.

Instead of just adding the values together, we can create a sequence of values by 
concatenating the iterations together. Starting with data and working down to a base 
case is called "recursion", while starting from a base case and building up a data 
structure is "corecursion".

If we want a sequence of values that represents the Fibonacci numbers, we can use

```{r}
fibs <- function(n) {
  if (n == 0) return(0)
  if (n == 1) return(1)
  prev <- fibs(n - 1)
  c(prev, sum(tail(prev, 2)))
}

fibs(10)
```

What's blown my mind recently is the concept of _infinite_ data structures. If I 
defined some function that, instead of working _down_ to a base case, just kept 
expanding, say, by concatenating with a _larger_ number (corecursion), such as

```{r}
inf_series <- function(x) {
  c(x, inf_series(x + 1))
}
```

Then, if I tried to evaluate `inf_series(5)` this would produce

```r
c(5, inf_series(6))
```

which would expand to

```r
c(5, 6, inf_series(7))
```

and so on... forever. Of course, R can't keep going forever

```{r, eval = FALSE}
inf_series(5)
```
```{r, eval = FALSE, class.source = "bg-danger"}
Error: C stack usage  7971732 is too close to the limit
```

This error comes about because each function in R is a "closure" which "encloses" 
its environment. The way R keeps track of that (and where it needs to return after 
returning from a function) is by adding "stack frames" each time it dives deeper 
into a function calling a function. We've asked it to add infinity of these, so 
at some point it says "too many".

Okay, so, not possible, right?

In Haskell I can define

```haskell
λ> fibs = 0 : scanl (+) 1 fibs
```

which is a concatenation (`:`) of `0` with the result of `scanl (+) 1 fibs`. To explain 
that part a little: [`scanl`](https://hackage.haskell.org/package/base-4.18.0.0/docs/Prelude.html#v:scanl) is similar to `reduce` in that it takes a starting value, a vector, and a 
binary operator, but rather than reducing the vector to a value, it creates a new 
vector with the successively reduced values. In this example, the values `1..5` are 
successively added (`+`) to `0`, so the second entry is `0+1=1`, the next is `1+2=3`, the 
next is `3+3=6`, then `6+4=10`, then `10+5=15`

```haskell
λ> scanl (+) 0 [1..5]
[0,1,3,6,10,15]
```

In the Fibonacci case, the operator is still addition, but the starting value is `1` 
and the vector is ... the entire vector we're defining. Writing out some of the
terms makes this easier to understand

```haskell
λ> scanl (+) 1 [0, 1, 1, 2, 3, 5, 8]
[1,1,2,3,5,8,13,21]
```

The first terms in the sequence, after concatenating with `0` will be 

```haskell
λ> [0, 1, 1, 2, 3, 5, 8]
```

so, by defining `fibs` in _terms of_ `fibs`, the sequence can go on forever. So, 
what if you try to print this? In `GHCI`, the output will just stream forever, which 
isn't particularly useful. Instead, we can ask for some number of values, say, ten

```haskell
λ> take 10 fibs
[0,1,1,2,3,5,8,13,21,34]
```

Due to the laziness of Haskell, nothing is computed until it's needed, so asking 
for any number of values is fast, despite the list being "infinite"

```haskell
λ> take 100 fibs
[0,1,1,2,3,5,8,13,21,34,55,89,144,233,377,610,987,1597,2584,4181,6765,10946,
 17711,28657,46368,75025,121393,196418,317811,514229,832040,1346269,2178309,
 3524578,5702887,9227465,14930352,24157817,39088169,63245986,102334155,
 165580141,267914296,433494437,701408733,1134903170,1836311903,2971215073,
 4807526976,7778742049,12586269025,20365011074,32951280099,53316291173,
 86267571272,139583862445,225851433717,365435296162,591286729879,956722026041,
 1548008755920,2504730781961,4052739537881,6557470319842,10610209857723,
 17167680177565,27777890035288,44945570212853,72723460248141,117669030460994,
 190392490709135,308061521170129,498454011879264,806515533049393,
 1304969544928657,2111485077978050,3416454622906707,5527939700884757,
 8944394323791464,14472334024676221,23416728348467685,37889062373143906,
 61305790721611591,99194853094755497,160500643816367088,259695496911122585,
 420196140727489673,679891637638612258,1100087778366101931,1779979416004714189,
 2880067194370816120,4660046610375530309,7540113804746346429,12200160415121876738,
 19740274219868223167,31940434634990099905,51680708854858323072,
 83621143489848422977,135301852344706746049,218922995834555169026]
```

The idea of taking some values from an iterator shows up in other languages. 

In Rust, I can create an infinite iterator of the value `1`

```rust
use std::iter;
let ones = iter::repeat(1);
```

and I can take some number of these, say, five, collected into a vector

```rust
ones.take(5).collect::<Vec<_>>()
```
```{rust, eval = FALSE, class.source = "bg-success"}
[1, 1, 1, 1, 1]
```

Python also has a notion of infinite sequences, and they're likely even more common.

When you work with a (regular, finite) list of values in a range, you get a `range` object

```{python}
numbers = range(1, 9)
numbers
type(numbers)
```

You can convert this to a regular list with

```{python}
list(numbers)
type(list(numbers))
```

If you `filter` the numbers (which works on a `range` or a `list`) you get a `filter` object

```{python}
even_numbers = filter(lambda x: x % 2 == 0, numbers)
even_numbers
type(even_numbers)
```

which you can also convert to a list to see the values

```{python}
list(even_numbers)
```

But, you can use this as an _iterable_, so you can get the 'next' value as 
many times as you need (defined here again to restart the iterator)

```{python}
even_numbers = filter(lambda x: x % 2 == 0, numbers)
next(even_numbers)
next(even_numbers)
next(even_numbers)
next(even_numbers)
```

until there's none left

```{python, error = TRUE}
next(even_numbers)
```

As before, you can convert these to a fixed-length list, if desired

```{python}
list(filter(lambda x: x % 2 == 0, numbers))
```

Still with me? Great. We can create an _infinite_ list, if we want to, because 
it isn't evaluated until we ask for elements

```{python}
def infinitenumbers():
    count = 0
    while True:
        yield count
        count += 1

nums = infinitenumbers()        

nums
type(nums)
```

This is a _generator_ which means it's capable of generating values. We can ask 
for as many as we want, now

```{python}
next(nums)
next(nums)
next(nums)
next(nums)
next(nums)
```

If we want to convert some number of these to a list, we need a new function, roughly 
the equivalent of Haskell's `take`, in order to extract these

```{python}
from itertools import islice

nums = infinitenumbers()        

list(islice(nums, 10))
```

and as before, if we ask for more _now_, we get the next batch

```{python}
list(islice(nums, 10))
```

So, can we do this at all in R? We can't build an infinite data structure _per se_, 
but what if we use a recursive definition and just... stop when it's recursed enough 
times?

I initially thought about hacking into functions with `body()` to keep track of this, 
but we've already discussed something we can use - the stack frames! R keeps track of 
how deep it's recursed using these, and we can access that information with `sys.calls()`, 
the help for which describes 

```
.GlobalEnv is given number 0 in the list of frames. Each subsequent function evaluation increases the frame stack by 1.
```

So, we can tell from inside a recursive function how deeply nested we currently are. 

I wrote the following helper, named for the Haskell inspiration

```{r}
take <- function(f, n, x = 1) {
  current_depth <- length(sys.calls())  # Subtract 1 to exclude the current call,
                                        # but add 1 to start at 1
  
  # uncomment this line to watch the magic happen!
  # cat("Current Stack Depth: ", current_depth, "\n")
 
  if (current_depth >= n) {
    return(f(x))
  } else {
    return(c(f(x), take(f, n, x + 1)))
  }
}
```

this checks `length(sys.calls())` which starts at 1 the first time it's called, and adds
1 every time we go deeper. So long as we haven't reached the requested depth, it 
combines the passed-in function evaluated at $x + i$ with a new evaluation one level deeper.

When that reaches the requested depth, it returns the evaluated function, bubbling up the
returned values so that we end up with a vector of $n$ values

$$
[f(x), f(x+1), f(x+2), f(x+3), \dots, f(x+n-1)]
$$

Neat idea, but does it work? We can't pass in an actual infinite data structure, 
but we _can_ pass a function that defines one

A (trivial) function that produces a number at each value of `x` is

```{r}
numbers <- function(x) {
  x
}
```

If we pretend that's a generator for every number, we can "take" some values from it

```{r, eval = FALSE}
take(numbers, 5)
```
```{r, eval = FALSE, class.source = "bg-success"}
 [1] 1 2 3 4 5
```

```{r, eval = FALSE}
take(numbers, 10)
```
```{r, eval = FALSE, class.source = "bg-success"}
 [1]  1  2  3  4  5  6  7  8  9 10
```

A more complicated recipe for an infinite list of numbers could be

```{r, eval = FALSE}
g <- function(x) {
  x + 1
}

take(g, 7)
```
```{r, eval = FALSE, class.source = "bg-success"}
[1] 2 3 4 5 6 7 8
```

```{r, eval = FALSE}
take(g, 10)
```
```{r, eval = FALSE, class.source = "bg-success"}
 [1]  2  3  4  5  6  7  8  9 10 11
```


Dealing with non-sequential numbers might be trickier... what if we want all the even 
numbers?

```{r, eval = FALSE}
evens <- function(x) {
  if(x %% 2 == 0) x else NULL
}

take(evens, 10)
```
```{r, eval = FALSE, class.source = "bg-success"}
 [1]  2  4  6  8 10
```

No, that only checks the first 10 numbers. Instead, 

```{r, eval = FALSE}
evens <- function(x) {
  x * 2
}

take(evens, 7)
```
```{r, eval = FALSE, class.source = "bg-success"}
 [1]  2  4  6  8 10 12 14
```
```{r, eval = FALSE}
take(evens, 10)
```
```{r, eval = FALSE, class.source = "bg-success"}
 [1]  2  4  6  8 10 12 14 16 18 20
```

What about our original example? 

```{r, eval = FALSE}
fibs <- function(x) {
  if (x == 0) return(0)
  if (x == 1) return(1)
  fibs(x - 1) + fibs(x - 2)
}

# 10 values, starting at 0
take(fibs, 10, x = 0)
```
```{r, eval = FALSE, class.source = "bg-success"}
 [1]  0  1  1  2  3  5  8 13 21 34
```
```{r, eval = FALSE}
# 10 values, starting at 1
take(fibs, 10, x = 1)
```
```{r, eval = FALSE, class.source = "bg-success"}
 [1]  1  1  2  3  5  8 13 21 34 55
```

Or, if we want 15 values, starting at the 10th

```{r, eval = FALSE}
take(fibs, 15, x = 10)
```
```{r, eval = FALSE, class.source = "bg-success"}
 [1]    55    89   144   233   377   610   987  1597  2584  4181  6765 10946 17711 28657 46368
```

I'd say that's working quite nicely!!!

Some caveats to keep in mind, though... 

Since we're relying on a count of stack frames on top of `.GlobalEnv`, this `take()` 
implementation won't work nicely inside another function. In fact, since 
{knitr} is already a few functions deep, it also doesn't work in an Rmd file (including 
this blog which is Rmd via {blogdown}). Not for use in production, but a fun
exercise to figure it out at all.

Is there a better way to achieve this `take()` functionality? Where do you use 
infinite iterators/generators in R or another language? Spot an improvement that 
I can make? I can be found on [Mastodon](https://fosstodon.org/@jonocarroll) or use the comments below.

<br />
<details>
  <summary>
    <tt>devtools::session_info()</tt>
  </summary>
```{r sessionInfo, echo = FALSE}
devtools::session_info()
```
</details>
<br />
