---
title: Iterative Square Root
author: Jonathan Carroll
date: '2024-05-29'
slug: iterative-square-root
categories:
  - rstats
  - APL
  - go
tags:
  - rstats
  - APL
  - go
type: ''
subtitle: ''
image: ''
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
  class.output  = "bg-success",
  class.message = "bg-info text-info",
  class.warning = "bg-warning text-warning",
  class.error   = "bg-danger text-danger"
)
```

I saw a toot celebrating a short, clean implementation of a square root 
finding algorithm and wanted to dig a bit deeper into how it works, with a 
diversion into some APL.

<!--more-->

I saw a toot celebrating a short, clean implementation of a square root 
finding algorithm and wanted to dig a bit deeper into how it works, with a 
diversion into some APL.

[This](https://hachyderm.io/@jimgar/112521236754983554) was the toot from Jim 
Gardner

> Doubly pleased with myself. 
> 
> Been doing the Tour of Go. Got to the section where you make a square root function, which should return once the calculated value stops changing. Struggled for ages. Trimmed and trimmed. Until finally... this!
>
> The calculation for z was given, and I don't understand it at all. But I don't care. It was a total mess when I started and has turned out quite neat. I'm very satisfied. 
>
> But why "doubly pleased"? Because I've been solely using Neovim so far for Go!! 

```{r, eval = FALSE}
package main 

import ( 
	"fmt" 
) 

func Sqrt(x float64) float64 {
	z := 1.0 
	for { 
		y := z 
		z -= (z*z - x) / (2 * z) 
		
		if z == y { 
			return z 
		} 
	}
} 

func main() { 
	fmt.Println(Sqrt(16)) 
} 
```

It's a nice, not-too-complicated algorithm to play with, and I agree it's hard to 
see _why_ it works for this application, so I thought it would be neat to walk 
through that.

What we're trying to solve here is the function $y = x^2$ which we could write as 
$f(x) = x^2 - y$ for which we want the value $x$ where $f(x) = 0$.

[Newton's Method](https://en.wikipedia.org/wiki/Newton%27s_method) is an iterative 
method for solving equations of this type (not all equations, mind you - I have an 
entire chapter of my PhD thesis discussing exactly _why_ it can't be used to 
solve the equations I was solving that my supervisor insisted it could). It works 
by using the slope (derivative) at a point to guide towards a solution. The formula 
for the updated value $x_{n+1}$ given some guess $x_n$ is 

$$
x_{n+1} = x_n - \frac{f(x_n)}{f'(x_n)}
$$
where $f'(x)$ is the derivative of the function $f$ at the point $x$. For $f(x) = x^2 - y$
the derivative is $f'(x) = 2x$ so we can substitute this and $f(x)$ into the above formula

$$
x_{n+1}​=x_n​−\frac{x_n^2-y}{2x_n}
$$
This is what the Go code calculates; given an initial guess of $x_n = 1$ it calculates the next value as

```
x = x - (x*x - y) / (2 * x) 
```

where, here, `y` is the value we're finding the square root of.

In R this could be written as

```r
SQRT <- function(x) {
  z <- 1
  while (TRUE) {
    y <- z
    z <- z - (z*z - x)/(2*z)
    if (abs(y -z) < 0.0001) return(z)
  }
}
```

(since `base::sqrt` is already defined) where I've used a tolerance rather than
relying on exact numerical equality. The `while(TRUE)` construct is equivalent
to Go's `for {}` syntax; an infinite loop.

R actually has another way to write that which is even closer; `repeat {}`

```r
SQRT <- function(x) {
  z <- 1
  repeat {
    y <- z
    z <- z - (z*z - x)/(2*z)
    if (abs(y -z) < 0.0001) return(z)
  }
}
```

One might notice that this approach requires essentially squaring a value, which 
is hardly expensive, but we can simplify and cancel out $x_n$, so

$$
x_{n+1}​=\frac{x_n-\frac{y}{x_n}}{2}
$$
in which case we have

```r
SQRT <- function(x) {
  z <- 1
  repeat {
    y <- z
    z <- (z + x/z)/2
    if (abs(y -z) < 0.0001) return(z)
  }
}
```

One of the reasons I wanted to dig into this was the fact that it's a convergence... 

In APL the power operator (`⍣` [aplwi](https://aplwiki.com/wiki/Power_(operator)))
applies a function some specified number of times, so 

```apl
    f ⍣n x
```

applies `f` to `x` `n` times, i.e. `(f⍣3)x` produces `f(f(f(x)))`.

It can also be used as `⍣=` where it will continue to apply the function until
the output no longer changes (is equal). A classic example is the 
[golden ratio](https://en.wikipedia.org/wiki/Golden_ratio); take the reciprocal
then add 1 until it converges, i.e. 

$$
x_{n+1} = 1+\frac{1}{x_n}
$$ 

which you can try for yourself 
[here](https://tryapl.org/?clear&q=1%2B%E2%88%98%C3%B7%E2%8D%A3%3D1&run)

```apl
    1+∘÷⍣=1
1.618033989
```

In this, `+∘÷` is the ([tacit](https://aplwiki.com/wiki/Tacit)) function created by 
composing (`∘`) 'addition of 1' (`1+`, a partial application of a function) and 
'reciprocal' (`÷`), which is iterated until it no longer changes (within 
machine precision).

Iterating until convergence is exactly what we want, since we're looking for the 
value satisfying 

$$
x_n = x_{n+1}​=\frac{x_n-\frac{y}{x_n}}{2}
$$
APL uses `⍵` as the right argument placeholder and `⍺` as the left, so the 
function we want to apply repeatedly to the right argument is

```apl
    {⍵-(((⍵×⍵)-⍺)÷(2×⍵))}
```

If we provide `1` as the right argument (the start value) and `16` as the left 
argument, we get

```apl
    16{⍵-(((⍵×⍵)-⍺)÷(2×⍵))}⍣=1
4
```

You can try this out yourself at [tryapl.org](https://tryapl.org/?clear&q=16%7B%E2%8D%B5-(((%E2%8D%B5%C3%97%E2%8D%B5)-%E2%8D%BA)%C3%B7(2%C3%97%E2%8D%B5))%7D%E2%8D%A3%3D1&run) 
(link should load that expression).

We can turn that into a function, once again using the argument placeholder

```apl
    sqrt←{⍵{⍵-(((⍵×⍵)-⍺)÷(2×⍵))}⍣=1}
    sqrt 25
5
    sqrt 81
9
```

Taking the simplification above, we can write this a bit shorter as

```apl
      sqrt←{⍵{(⍵+(⍺÷⍵))÷2}⍣=1}
      sqrt 144
12
```

As clean as the Go code looks, I think there's a certain beauty to being able 
to write this in just 20 characters. It's not for everyone, I get that.

I love these opportunities to learn a bit more about languages.

If you have comments, suggestions, or improvements, as always, feel free to use
the comment section below, or hit me up on
[Mastodon](https://fosstodon.org/@jonocarroll).

<br />
<details>
  <summary>
    <tt>devtools::session_info()</tt>
  </summary>
```{r sessionInfo, echo = FALSE}
devtools::session_info()
```
</details>
<br />
