---
title: Haskell IS a Great Language for Data Science
author: Jonathan Carroll
date: '2025-11-20'
categories:
  - haskell
  - rstats
tags:
  - haskell
  - rstats
slug: haskell-is-a-great-language-for-data-science
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
  class.output  = "bg-success",
  class.message = "bg-info text-info",
  class.warning = "bg-warning text-warning",
  class.error   = "bg-danger text-danger"
)
knitr::opts_chunk$set(fig.path = "") 
```

I've been learning Haskell for a few years now and I am really liking a lot of 
the features, not least the strong typing and functional approach. I thought it 
was lacking some of the things I missed from R until I found the dataHaskell 
project. In this post I'll demonstrate some of the features and explain why I 
think it makes for a good (great?) data science language.

<!--more-->

I've been learning Haskell for a few years now and I am really liking a lot of
the features, not least the strong typing and functional approach. I thought it
was lacking some of the things I missed from R until I found the
[dataHaskell](https://www.datahaskell.org/) project.

There have been several attempts recently to enhance R with some strong types, e.g. 
[vapour](https://vapour.run/) and [typr](https://github.com/we-data-ch/typr), but 
while these try to bend R towards types, perhaps an all-in solution makes more sense.

In this post I'll demonstrate some of the features and explain why I think it
makes for a good (great?) data science language.

I've posted [more than a handful of times](https://jcarroll.com.au/tags/haskell/) 
about Haskell but maybe not so much the benefits of a real-world usage, more 
toy problems (e.g. I did a lot of Advent of Code using it last year). I've been 
working towards using it more, and even managed to get a custom {knitr} engine 
working - here's the special sauce that makes a ` ```{haskell} ` block work:

```{r}
knitr::knit_engines$set(haskell = function(options) {
  code <- options$code
  codefile <- tempfile(fileext = ".hs")
  on.exit(file.remove(codefile))
  writeLines(c(":script dataframe", "", code), con = codefile)
  out  <- system2(
    file.path(path.expand('~'), '.ghcup/bin/ghc'),
    c('-e',"':script ", codefile, "'"),
    stdout = TRUE
  )

  knitr::engine_output(options, code, out)
})
```

This writes the lines of code to a temporary file, prepended with some
configuration options, then runs essentially `ghc -e ':script file.txt'` and
deletes the temporary file. It also adds a package just in case that's not
exposed to the session. The result is then shown in the code block, so this is a
"live" output

```{haskell}
map (+5) [2..8]
```

Neat, right?

So, what would one use this for?

I saw [this (follow-up) post](https://blog.genesmindsmachines.com/p/python-is-not-a-great-language-for-2e0)
from Claus Wilke about Python not being a great language for data science and
while I concur with the points made there, I do believe some of them are
personal preference. I'm a proponent of "use the tools you're comfortable with"
and I can't argue with however many thousands of data scientists are
successfully using Python to do data science.

The point about "what makes for a good data science language" made me pause to 
think and I came to the conclusion that Haskell actually ticks the boxes, at least 
with the dataHaskell ecosystem and the dataframe package. What follows is not to 
be taken as a pile-on against Python or even a complaint about R, but rather 
something in the style of "if you like that, check this out!"

Lots of languages seem to have some sort of dataframe these days - thanks R! - e.g. 
Python has Pandas/Polars, Julia has DataFrames.jl, even Kotlin has a DataFrame. 
Haskell does, too with [dataframe](https://dataframe.readthedocs.io/) and I've 
been learning how to use this recently.

The points made in Claus' post were that the features which make R a better language
for data science over Python are (paraphrasing):

- call-by-value semantics (non-mutability)
- built-in missing values
- built-in vectorization
- non-standard evaluation (NSE)

Let's look at how Haskell deals with each of these.

## Non-mutability

Claus details how Python's call-by-reference semantics enables one to modify 
variables unintentionally, since they're scoped across functions. Haskell certainly 
doesn't have this problem - everything is immutable, and functions are "pure" (no 
side-effects, though you can interact with typed side-effect 'instructions'). If you 
want to "do" anything to a data object you pass it into a function and get a new 
object out. There's no risk of accidentally modifying a variable, but of course 
the downside of this is that you _can't_ do so without a function. While in R it's 
straightforward to do 

```{r}
a <- c(2, 9, 6)
a[2] <- 4
a
```

in Haskell you'd need to break the vector apart and stitch the new value inside

```{haskell}
a = [2, 9, 6]

:{
updateSecond (x:_:z) y = x : y : z
updateSecond xs _ = xs
:}

updateSecond a 4
```

[Multi-line statements](https://sanj.ink/posts/2018-08-09-defining-a-multiline-function-in-haskell-using-ghci.html) need to be enclosed between `:{` and `:}`.

No risk of accidentally writing that, I'm sure.

## Built-in missing values

This is somewhere that Haskell shines - if you want a value that _might_ not be
available in R you use an `NA` (which is a shorthand for whichever flavour/class 
you actually want, e.g. `NA_character_`). Using one of these in any mathematical 
calculation 'poisons' it and returns `NA`, e.g.

```{r}
sum(1, NA, 3)
```

To avoid this, most functions offer a `na.rm` argument

```{r}
sum(1, NA, 3, na.rm = TRUE)
```

What's happening here is that R encodes a value that is _maybe_ missing. Haskell 
formalises this into the `Data.Maybe` package and you have to be explicit in 
dealing with a missing value (`Nothing`) or a definitely-not-missing value (`Just x`)

```{haskell}
non_missing = [1, 2, 3, 4]
has_missing = [Just 1,Just 2,Nothing,Just 4]

:t non_missing
:t has_missing
```

where we see that `has_missing` is a `Maybe` type. 

```{haskell, echo = -c(1, 2)}
non_missing = [1, 2, 3, 4]
has_missing = [Just 1,Just 2,Nothing,Just 4]
sum non_missing
```

You can't just `sum` the latter

```haskell
sum has_missing
```
```{r, eval = FALSE, class.source = "bg-danger"}
s:7:1: error: [GHC-39999]
    • No instance for ‘Num (Maybe Integer)’ arising from a use of ‘it’
    • In the first argument of ‘print’, namely ‘it’
      In a stmt of an interactive GHCi command: print it
  |
7 | sum has_missing
  | ^^^^^^^^^^^^^^^
```

you need to remove any `Nothing` first, then most likely 'unwrap' from the 
`Maybe` context

```{haskell, echo = -c(1)}
has_missing = [Just 1,Just 2,Nothing,Just 4]
import Data.Maybe
sum $ map fromJust $ filter isJust has_missing
```

or alternatively

```{haskell, echo = -c(1,2)}
has_missing = [Just 1,Just 2,Nothing,Just 4]
import Data.Maybe
sum (mapMaybe id has_missing)
```

or you can get fancy

```{haskell, echo = -c(1)}
has_missing = [Just 1,Just 2,Nothing,Just 4]
sum [x | Just x <- has_missing]
```

The point is that you _have to_ deal with the missingness if it's there. What this
also means is that if you have a `Double` column, it _does NOT_ have missing values, 
so you can safely sum those values (plus get all sorts of performance benefits from 
the compiler because it, too, knows there's no missing values).

For Claus' example, we can produce a proper `Nothing` at the end of this calculation

```{haskell}
import Data.Maybe 
x = [Just 1,Just 2,Nothing,Just 4,Just 5]
fmap (fmap (> 3)) x
```

## Built-in vectorisation

Haskell is NOT an array language, so sure, it doesn't have vectorisation built-in, 
but it's worth noting that at the end of Claus' post he details some limitations 
of R and acknowledges that "R does not have any scalar data types". Haskell has 
scalars, vectors, and arrays, and you need to be specific when you want to iterate 
over those - the "type" of a variable includes the dimensionality, so `Double` is 
not the same as `[Double]` (a list of doubles). 

Since Haskell is a functional programming language it has every type of `map` you 
could want, including specialities for monads and applicatives. While this means 
you _do_ need to write `map` when you want to iterate, it also means you're never 
surprised that there was more than one value there.

What's more, because it's a compiled language, the compiler can optimise all sorts 
of vector operations. One example is using "fusion" to combine a `filter` and a 
`map` such that the 
[intermediate vector isn't actually used at all](https://www.tweag.io/blog/2017-10-12-vector-package/#array-fusion).

## Non-standard evaluation (NSE)

This is where the fun really starts - the dataframe package in Haskell adds 
the sort of slicing and dicing you're probably familiar with 

```{haskell}
df <- D.readParquet "iris.parquet"

D.nRows df
```

(don't be fooled by that `<-` - that's Haskell's way of doing something that reaches 
outside of the CPU, e.g. to the disk to read a file). 

```{haskell}
iris <- D.readParquet "iris.parquet"

:{
iris |> 
  D.filterWhere (F.col @T.Text "variety" .== "Setosa") |> 
  D.filterWhere (F.col @Double "sepal.length" .> 5.4)
:}
```

but goes one step further via template haskell... you can expose the columns as
variables (admittedly, in the wider scope) so this works

```{haskell}
iris <- D.readParquet "iris.parquet"

-- make columns available as expressions
:exposeColumns iris

:{
iris |> 
  D.derive "ratio" (sepal_width / sepal_length) |>
  D.take 5
:}
```

(explain output and redirection)

<br />
<details>
  <summary>
    <tt>devtools::session_info()</tt>
  </summary>
```{r sessionInfo, echo = FALSE}
devtools::session_info()
```
</details>
<br />
