---
title: Haskell IS a Great Language for Data Science
author: Jonathan Carroll
date: '2025-11-20'
categories:
  - haskell
  - rstats
tags:
  - haskell
  - rstats
slug: haskell-is-a-great-language-for-data-science
---

```{r, setup, include = FALSE}
options(crayon.enabled = TRUE)
ansi_aware_handler <- function(x, options) {
  paste0(
    "<pre class=\"bg-success\"><code>",
    fansi::sgr_to_html(x = x, warn = FALSE, term.cap = "256"),
    "</code></pre>"
  )
}
knitr::knit_hooks$set(output = ansi_aware_handler)
knitr::opts_chunk$set(
  class.output  = "bg-success",
  class.message = "bg-info text-info",
  class.warning = "bg-warning text-warning",
  class.error   = "bg-danger text-danger"
)
knitr::opts_chunk$set(fig.path = "") 
```

I've been learning Haskell for a few years now and I am really liking a lot of 
the features, not least the strong typing and functional approach. I thought it 
was lacking some of the things I missed from R until I found the dataHaskell 
project. In this post I'll demonstrate some of the features and explain why I 
think it makes for a good (great?) data science language.

<!--more-->

I've been learning Haskell for a few years now and I am really liking a lot of
the features, not least the strong typing and functional approach. I thought it
was lacking some of the things I missed from R until I found the
[dataHaskell](https://www.datahaskell.org/) project.

There have been several attempts recently to enhance R with some strong types, e.g. 
[vapour](https://vapour.run/), [typr](https://github.com/we-data-ch/typr), using 
[{rlang}'s checks](https://josiahparry.com/posts/2024-06-30-type-safety/), and even 
discussions about implementations at the core level e.g. 
[in September 2025](https://stat.ethz.ch/pipermail/r-devel/2025-September/084164.html) 
continued [in November 2025](https://stat.ethz.ch/pipermail/r-devel/2025-November/084223.html).
While these try to bend R towards types, perhaps an all-in solution makes more sense.

In this post I'll demonstrate some of the features and explain why I think it
makes for a good (great?) data science language.

I've posted [more than a handful of times](https://jcarroll.com.au/tags/haskell/) 
about Haskell but maybe not so much the benefits of a real-world usage, more 
toy problems (e.g. I did a lot of Advent of Code using it last year). I've been 
working towards using it more, and even managed to get a custom {knitr} engine 
working - here's the special sauce that makes a ` ```{haskell} ` block work:

```{r}
knitr::knit_engines$set(haskell = function(options) {
  code <- options$code
  codefile <- tempfile(fileext = ".hs")
  codefile_brace <- tempfile(fileext = ".hs")
  on.exit(file.remove(codefile, codefile_brace))
  writeLines(c(":script dataframe", "", code), con = codefile)
  system2('hscript', codefile, stdout = codefile_brace)
  out  <- system2(
    file.path(path.expand('~'), '.ghcup/bin/ghc'),
    c('-e',"':script ", codefile_brace, "'"),
    stdout = TRUE
  )

  knitr::engine_output(options, code, out)
})
```

This writes the lines of code to a temporary file, prepended with some
configuration options, then runs essentially `ghc -e ':script file.txt'` and
deletes the temporary file. For the purposes of making cleaner code blocks, the 
code detours through an [`awk` script](https://github.com/DataHaskell/datahaskell-starter/blob/main/hscript) 
which inserts some `:{` blocks around 
[multi-line statements](https://sanj.ink/posts/2018-08-09-defining-a-multiline-function-in-haskell-using-ghci.html), 
helping to reproduce how these look in a Jupyter notebook. 
The result is then shown in the code block, so this is a "live" output

```{haskell}
map (+5) [2..8]
```

Neat, right?

Because I'm treating each code block as an independent script, it means there is 
_some_ repetition between blocks. I'll hide that away with some judicious `echo` 
options where necessary, but otherwise each block should be able to be run 
as a 'script' with the right pre-processing.

<a id="syntax"></a>

## A Brief intro to Haskell Syntax

Haskell _is_ a bit different if you've only ever seen R or Python, but it doesn't 
take too much effort to understand what's going on. Firstly, while parentheses are 
used for function calls in R, a space is used in Haskell, so instead of `sum(x)` 
you use `sum x`. Parentheses are still used for grouping together combinations of 
things that need to be evaluated together.

Lists are a fundamental data type and are denoted by square brackets, e.g. `[3,4,5]` 
and they need to contain a single type. For a strongly typed language, that shouldn't 
come as a surprise. A single number might be of type `Double` and a list of these 
would be of type `[Double]`.

If you're worried that you've become too reliant on a piped workflow, fear not! 
Haskell has the familiar pipe operator

```{haskell}
[2,8,7,10,1,9,5,3,4,6] |>
  reverse |>
  take 5
```

with the important distinction that it passes the left side to the _end_ of the 
right side (not to the first argument) which flows cleaner given how Haskell 
functions are typically written, e.g.

```{haskell}
take 3 [1,2,3,4,5,6]

-- vs 

[1,2,3,4,5,6] |>
  take 3
```

The line in the middle there demonstrates that comments start with two hyphens `--`, 
or for multi-line comments, between `{-` and `-}`.

If you need to write a function (for which you use camelCase) you can annotate 
it with a definition, though the compiler can figure this out itself most of the 
time (plus it helps for readability). The way to do this is with one extra line 
above the implementation. If the type is generic, you can use a placeholder rather 
than a _specific_ type

```{haskell}
appendValueToList :: [a] -> a -> [a]
appendValueToList xs y = xs ++ [y]

appendValueToList [2,4,6] 8

appendValueToList ["f", "o", "o"] "t"
```

The period is used for function composition, i.e.

```{haskell}
import Data.List (sort)

(reverse . sort) [2,8,7,10,1,9,5,3,4,6]
```

applies a composed 'sort and reverse' operation to the list. The `import` is there 
because the 'base' library ("Prelude") doesn't have the `sort` function, so it's 
imported. There's actually a few of these which need to be imported to use the 
code I'm showing below, but it's inserted into the codeblocks via the `:script dataframe` 
line in the engine definition above. That calls out to an executable which runs 
the code block as if it was contained in a `main` function in a full program, 
which enables us to use `IO` operations inline, such as reading from files and 
printing results. That all gets a little trickier without this 'scripting' context, 
but I'm here to make the point that such a scripting context _works well_ for 
doing data science.

So, what would one use this for?

I saw [this (follow-up) post](https://blog.genesmindsmachines.com/p/python-is-not-a-great-language-for-2e0)
from Claus Wilke about Python not being a great language for data science and
while I concur with the points made there, I do believe some of them are
personal preference. I'm a proponent of "use the tools you're comfortable with"
and I can't argue with however many thousands of data scientists are
successfully using Python to do data science.

The point about "what makes for a good data science language" made me pause to 
think and I came to the conclusion that Haskell actually ticks the boxes, at least 
with the [dataHaskell ecosystem](https://www.datahaskell.org/blog/2025/11/11/welcome-to-datahaskell.html) 
and the dataframe package. What follows is not to 
be taken as a pile-on against Python or even a complaint about R, but rather 
something in the style of "if you like that, check this out!"

Lots of languages seem to have some sort of dataframe these days - thanks R! - e.g. 
Python has [Pandas](https://pandas.pydata.org/)/[Polars](https://docs.pola.rs/api/python/stable/reference/index.html), 
Julia has [DataFrames.jl](https://dataframes.juliadata.org/stable/), 
[even Kotlin has a DataFrame](https://kotlin.github.io/dataframe/home.html). 
Haskell does, too with [dataframe](https://dataframe.readthedocs.io/) and I've 
been learning how to use this recently.

The points made in Claus' post were that the features which make R a better language
for data science over Python are (paraphrasing):

- call-by-value semantics (non-mutability)
- built-in missing values
- built-in vectorization
- non-standard evaluation (NSE)

Let's look at how Haskell deals with each of these.

## Non-mutability

Claus details how Python's call-by-reference semantics enables one to modify 
variables unintentionally, since they're scoped across functions. Haskell certainly 
doesn't have this problem - everything is immutable, and functions are "pure" (no 
side-effects, though you can interact with typed side-effect 'instructions'). If you 
want to "do" anything to a data object you pass it into a function and get a new 
object out. There's no risk of accidentally modifying a variable, but of course 
the downside of this is that you _can't_ do so without a function. While in R it's 
straightforward to do 

```{r}
a <- c(2, 9, 6)
a[2] <- 4
a
```

in Haskell that sort of thing is off limits - you can use an operator to extract 
a value from a list (0-indexed), e.g. 

```{haskell}
a = [2,9,6]

a !! 1
```

but there's no way to assign the second element to some other value. Instead, 
you need to break the vector apart and stitch the new value inside

```{haskell}
a = [2,9,6]

updateSecond :: [a] -> a -> [a]
updateSecond (x:_:z) y = x : y : z
updateSecond xs _ = xs

updateSecond a 4
```

No risk of accidentally writing that, I'm sure. 

I've also included the type definition in this case which reads as "a function
which takes a list of some type `a` (`[a]`) and a single value of type `a` and
returns a list of that same type, `[a]`." FYI, this is one example where you may
need the definition to be enclosed between `:{` and `:}`, if you're running
interactively in `ghci`, but here I'm using a neat 
[pre-processing trick](https://github.com/DataHaskell/datahaskell-starter/blob/main/hscript).

A tick for _truly_ immutable data - the only way to alter a value is to operate 
on it with a function.

## Built-in missing values

This is somewhere that Haskell shines - if you want a value that _might_ not be
available in R you use an `NA` (which is a shorthand for whichever flavour/class 
you actually want, e.g. `NA_character_`). Using one of these in any mathematical 
calculation 'poisons' it and returns `NA`, e.g.

```{r}
sum(1, NA, 3)
```

To avoid this, most functions offer a `na.rm` argument

```{r}
sum(1, NA, 3, na.rm = TRUE)
```

What's happening here is that R encodes a value that is _maybe_ missing. Haskell 
formalises this into the `Data.Maybe` package and you have to be explicit in 
dealing with a missing value (`Nothing`) or a definitely-not-missing value (`Just x`)

```{haskell}
non_missing = [1, 2, 3, 4]
has_missing = [Just 1,Just 2,Nothing,Just 4]

:t non_missing
:t has_missing
```

where we see that `has_missing` is a `Maybe` type. 

```{haskell, echo = -c(1:3)}
non_missing = [1, 2, 3, 4]
has_missing = [Just 1,Just 2,Nothing,Just 4]

sum non_missing
```

You can't just `sum` the latter; it produces an error because it doesn't have a
function which can sum a `Maybe Integer`

```haskell
sum has_missing
```
```{r, eval = FALSE, class.source = "bg-danger"}
s:7:1: error: [GHC-39999]
    • No instance for ‘Num (Maybe Integer)’ arising from a use of ‘it’
    • In the first argument of ‘print’, namely ‘it’
      In a stmt of an interactive GHCi command: print it
  |
7 | sum has_missing
  | ^^^^^^^^^^^^^^^
```

you need to remove any `Nothing` first, then most likely 'unwrap' from the 
`Maybe` context

```{haskell, echo = -c(1:2)}
has_missing = [Just 1,Just 2,Nothing,Just 4]

import Data.Maybe

sum $ map fromJust $ filter isJust has_missing
```

or alternatively

```{haskell, echo = -c(1:4)}
has_missing = [Just 1,Just 2,Nothing,Just 4]

import Data.Maybe

sum (mapMaybe id has_missing)
```

or you can get fancy

```{haskell, echo = -c(1:2)}
has_missing = [Just 1,Just 2,Nothing,Just 4]

sum [x | Just x <- has_missing]
```

The point is that you _have to_ deal with the missingness if it's there. What this
also means is that if you have a `Double` column, it _does NOT_ have missing values, 
so you can safely sum those values (plus get all sorts of performance benefits from 
the compiler because it, too, knows there's no missing values).

For Claus' example, we can produce a proper `Nothing` at the end of this calculation

```{haskell, echo = -c(1:4)}
import Data.Maybe 

x = [Just 1,Just 2,Nothing,Just 4,Just 5]

fmap (fmap (> 3)) x
```

Another tick - proper missing values.

## Built-in vectorisation

Haskell is NOT an array language, so sure, it doesn't have vectorisation built-in, 
but it's worth noting that at the end of Claus' post he details some limitations 
of R and acknowledges that "R does not have any scalar data types". Haskell has 
scalars, vectors, and arrays, and you need to be specific when you want to iterate 
over those - the "type" of a variable includes the dimensionality, so `Double` is 
not the same as `[Double]` (a list of doubles). 

Since Haskell is a functional programming language it has every type of `map` you 
could want, including specialities for monads and applicatives. While this means 
you _do_ need to write `map` when you want to iterate, it also means you're never 
surprised that there was more than one value there.

What's more, because it's a compiled language, the compiler can optimise all sorts 
of vector operations. One example is using "fusion" to combine a `filter` and a 
`map` such that the 
[intermediate vector isn't actually used at all](https://www.tweag.io/blog/2017-10-12-vector-package/#array-fusion).

## Non-standard evaluation (NSE)

This is where the fun really starts - the [dataframe](https://dataframe.readthedocs.io/en/latest/) 
package from the dataHaskell ecosystem adds the sort of slicing and dicing
you're probably familiar with. Apart from general inspection of data frames

```{haskell}
df <- D.readParquet "iris.parquet"

D.describeColumns df
```

(don't be fooled by that `<-` - that's Haskell's way of doing something that reaches 
outside of the CPU, e.g. to the disk to read a file) we can use `D.dimensions` to get the 
overall shape, and more specific helpers like `D.nRows` and `D.nColumns` are available
which we can incorporate into e.g. text output

```{haskell}
import Text.Printf (printf)

df <- D.readParquet "iris.parquet"

D.dimensions df

printf "%d rows, %d columns" (D.nRows df) (D.nColumns df)
```

Many of the `dplyr`-esqe operations are available, with a lot of thought put into 
how these would interact with a strongly typed structure

```{haskell}
iris <- D.readParquet "iris.parquet"

iris |> 
  D.filterWhere (F.col @Text "variety" .== "Setosa") |> 
  D.filterWhere (F.col @Double "sepal.length" .> 5.4)
```

but dataframe goes one step further via template haskell... you can expose the
columns as variables (admittedly, in the wider scope) so this works

```{haskell}
iris <- D.readParquet "iris.parquet"

-- make columns available as expressions
:exposeColumns iris

iris |> 
  D.derive "sepal.ratio" (sepal_width / sepal_length) |>
  D.take 5 
```

What you _don't_ see here (because it's piped to `stderr`) is the info about the 
exposed columns

```
sepal_length :: Expr Double
sepal_width :: Expr Double
petal_length :: Expr Double
petal_width :: Expr Double
variety :: Expr Text
```

and it's worth noting that the dots/periods have been replaced by underscores. 
That's because in Haskell the period is used for composition, as described 
[above](#syntax).

Many verbs are supported, so we can do some more detailed transformations

```{haskell}
iris <- D.readParquet "iris.parquet"

:exposeColumns iris

iris |> 
  D.filterWhere ( sepal_width .> 2.6 ) |>
  D.groupBy ["variety"] |> 
  D.aggregate
      [ "n"       .= F.count petal_length
      , "sl_mean" .= F.mean sepal_length
      , "pl_mean" .= F.mean petal_length
      ]
```

and remember, because we _know_ there's no missing values in a column of `Double`s 
(not `Maybe Double`s) we can take averages without worrying about any `na.rm` 
complications. 

The NSE bit doesn't _quite_ work everywhere, but sometimes a string is just fine, e.g.

```{haskell, warning = TRUE}
iris <- D.readParquet "iris.parquet"

D.plotScatter "sepal.length" "sepal.width" iris
```

(following https://blog.djnavarro.net/posts/2021-04-18_pretty-little-clis/ to get 
the ANSI sequences to work in a code block). 

A more detailed comparison to `dplyr` is provided in 
[the dataframe documentation](https://dataframe.readthedocs.io/en/latest/coming_from_other_implementations.html#coming-from-dplyr).

So, NSE? Tick!

## Conclusion

I've hopefully demonstrated some of the power of a strongly typed language and 
a package focused on data science enabling the sort of functionality that an 
R (or Python) user might be looking for. I am hopeful that Haskell (and the 
dataHaskell ecosystem) can be a viable option for those of us wanting to do 
data science in a strongly typed language with a very clever compiler capable 
of making significant performance improvements.

If you're interested in dataHaskell then check out 
[this post](https://www.datahaskell.org/blog/2025/11/11/welcome-to-datahaskell.html) 
and consider taking it for a spin - we're working on reducing friction to get 
started via devcontainers and hosted notebook solutions, and are keen to hear from 
more data scientists about what they'd like the ecosystem to be able to support.

_I believe Haskell IS a great language for data science!_

As always, I can be found on
[Mastodon](https://fosstodon.org/@jonocarroll) and the comment section below.

<br />
<details>
  <summary>
    <tt>devtools::session_info()</tt>
  </summary>
```{r sessionInfo, echo = FALSE}
devtools::session_info()
```
</details>
<br />
